### 2021-09-17 products/manuscript/Manuscript.Rmd
Initial Commit
Updated Readme to reflect additions to meet Project Part 1 requirements.

---
title: "MADA Data Analysis Project"
subtitle: ""
author: Monica Chan
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

The structure below is a possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed.

# Summary/Abstract
_Write a summary of your project._  
Applying MADA learned techniques to a RT-PCR dataset to determine the effectiveness of sample enhancement strategies for environmental surface sampling of low bio-burden viruses.

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

### Surfaces Preparation  
Stainless-steel coupons (10x20in) were triple washed with detergent, rinsed with RO water, sprayed with 70% Ethanol and left to air dry. The dry coupons were wrapped in craft paper and autoclaved on a solid cycle.  Before use, the coupons are unwrapped while in a BSC

### Inoculum Preparation
A serial dilution of heat-inactivated SARS-CoV-2 (~10^6 copies/uL, ATCC Strain) and Viral Transport Media (CDC protocol) was created

### Inoculation of Surfaces

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

Data was self recorded over several experimental runs of an on going series. 
The current set has 379 observations and 23 variables. Additional data will be added into this set as experiments are completed throughout the course.
The data itself is a combination of observed processing variables and the resulting RT-PCR data. Copy data may eventually be included. It is currently calculated in an excel document, but may be reproduced in this project for best practice).
A copy of the data has been placed in the data folder of this repository.

## Questions/Hypotheses to be addressed

There are 3 major aims of this:

1. Do the sample enhancement strategies work?
2. How well do they work compared to each other?
3. Are there any environmental variables recorded that have a positive impact on detection?

Testing several sample enhancement strategies (Nanoparticles and Bait & Capture) for molecular detection of heat-inactivated SARS-CoV-2 from stainless-steel surfaces. The primary aim of this work is to determine if any of the enhancement strategies increase the concentration of RNA copies in a sample that will express itself as a lower Ct value from the Rt-PCR output. Lower Ct values equate to more copies being present in the sample meaning less time needed to confirm presence.

In the case the sample enhancements work, we would then compare how well they worked against an automated extraction method. The degree of "wellness" will be determined by how low the Ct value results are produced after a sample has going through a specific method.

In addition to the comparison of enhancement strategies the effects of the tool used for sampling (macrofoam or knit-fiber), time held, and other environmental variables will be explored any significant effects, with interest in positive impacts to detection.

# Methods and Results

_In most research papers, results and methods are separate. You can combine them here if you find it easier. You are also welcome to structure things such that those are separate sections._

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Exploratory analysis

_Use a combination of text/tables/figures to explore and describe your data. You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. FIgures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data._

_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables. Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones._

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

Table \@ref(tab:summarytable) shows a table summarizing the data.

```{r summarytable,  echo=FALSE}
resulttable=readRDS("../../results/summarytable.rds")
knitr::kable(resulttable, caption = 'Data summary table.')
```

Figure \@ref(fig:resultfigure) shows a scatterplot figure produced by one of the R scripts.

```{r resultfigure,  fig.cap='Analysis figure.', echo=FALSE}
knitr::include_graphics("../../results/resultfigure.png")
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.

```{r resulttable,  echo=FALSE}
resulttable=readRDS("../../results/resulttable.rds")
knitr::kable(resulttable, caption = 'Linear model fit table.')
```

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@Leek2015a] discusses types of analyses.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

# References

### 2021-10-09 products/Part_2/DataExploration.Rmd
Data Exploration Commit (07)

---
title: "Data Exploration"
subtitle: ""
author: Monica Chan
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::html_document2: 
    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: true
bibliography: ../references.bib
csl: ../apa.csl
---
# Data Exploration
Complete the data loading/cleaning/wrangling part, as well as having a mostly complete data exploration part.

## Data Loading  
_What the data is_

Compilation of RT-PCR results for a couple of probes (N1 & N2) to detect recovered heat-inactivated SARS-CoV-2 samples that have undergone sample enhancement procedures.
These data can be found in the data folder under raw_data

## Data Cleaning
Cleaned data is found under the data in processed_data folder

1. Compiled reports into a single csv.
2. Import the reports into R.
3. Check summary, head, and tail of data 
4. Select relevant parts of the data to form new data frame
5. Mutate classes of data for easier manipulation.
6. Renamed a value as it is the only value left. The Raw data had sevearl versions of it as the series of columns displayed a train of thought display of calculation.

* noted that there were additional NAs that would be generated when mutating Ct values into numeric. Which should be fine, but with how the reports are generated some levels recorded are noted as Undetermined or not reliably detected.
* Decide to Split the Data and make a companion UndeterminedData dataframe for additional analysis.

__Making the UndeterminedData__
see above but added filter() for Ct=="Undetermined" to pull all data with Undetermined as a reported value

## Data Wrangling

Started to play with analysisscript.
Outputs of the data play are in results.

####Run summaries

####Exploritory Figures

##### View the whole data set

![resultfigure](../../results/resultfigure.png)

##### Breakdown comparisons with the two methods being compared-- CDC and NTP had a comparitive input of 300uL

![resultfigure2](../../results/resultfigure2.png)
Interesting-- why is there such a group difference in Surface Sampling experiements versus Spike
Oh makes sense it's SD- the serial dilution. We expect the SD to be like the standard curve and the recovery should have loss.

Deeper dive into looking at the Surface Sampling Experiment, as it has some differences and is new-er data. Also includes data with Temp & RH:

##### Breakdown comparisons more with only the Surface Sample experiment

![SS300](../../results/SS300.png)

### 2021-10-09 products/manuscript/Manuscript.Rmd
Data Exploration Commit (07)

@@
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+
+Data comes out of different reports. This report is added to a supplemental report gathered while the samples were processed and added into one giant compilation then read into R.
+
+processingscript.R was modified to do initial cleaning and preparation of the dataset for further exploration.Data is also saved.
 

### 2021-10-09 code/analysis_code/analysisscript.R

@@ -12,16 +12,18 @@ library(here) #for data loading/saving
 #path to data
 #note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed_data","processeddata.rds")
+data_location2 <- here::here("data","processed_data","UndeterminedData.rds")
+
+
 
 #load data. 
 mydata <- readRDS(data_location)
-
+Un_mydata <- readRDS(data_location2)
 ######################################
 #Data exploration/description
 ######################################
-#I'm using basic R commands here.
-#Lots of good packages exist to do more.
-#For instance check out the tableone or skimr packages
+
+
 
 #summarize data 
 mysummary = summary(mydata)
@@ -40,7 +42,12 @@ saveRDS(summary_df, file = summarytable_file)
 
 #make a scatterplot of data
 #we also add a linear regression line to it
-p1 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
+p1 <- mydata %>% 
+  filter(Method!="PCR")%>%
+  ggplot(aes(x=Dilution.Factor, y=Ct)) + 
+  geom_jitter(aes(color=Method)) + 
+  geom_smooth(method='lm')+
+  facet_grid(Method~Type)
 
 #look at figure
 plot(p1)
@@ -54,7 +61,7 @@ ggsave(filename = figure_file, plot=p1)
 ######################################
 
 # fit linear model
-lmfit <- lm(Weight ~ Height, mydata)  
+lmfit <- lm(Dilution.Factor ~ Ct, mydata)  
 
 # place results from fit into a data frame with the tidy function
 lmtable <- broom::tidy(lmfit)
@@ -66,4 +73,92 @@ print(lmtable)
 table_file = here("results", "resulttable.rds")
 saveRDS(lmtable, file = table_file)
 
-  
+####################################################################
+#Data Exploration More:
+######################
+
+#Summary table based on 
+SummaryTableAll<-mydata%>%
+  group_by(EXP,Target, Dilution.Factor, Method, Type)%>%
+  summarise(n=n(),
+            Mean= mean(Ct),
+            SD= sd(Ct))
+
+print(summary(SummaryTableAll))
+
+#attempt to see if i can get a better summary table?
+summary_df2 = data.frame(do.call(cbind, lapply(SummaryTableAll, summary)))
[…]

### 2021-10-09 code/processing_code/processingscript.R

@@ -8,10 +8,10 @@
 library(readxl) #for loading Excel files
 library(dplyr) #for data processing
 library(here) #to set paths
-
+library(stringr)
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw_data","exampledata.xlsx")
+data_location <- here::here("data","raw_data","2021.10.08_Spike2.csv")
 
 #load data. 
 #note that for functions that come from specific packages (instead of base R)
@@ -19,42 +19,34 @@ data_location <- here::here("data","raw_data","exampledata.xlsx")
 #package::function() that's not required one could just call the function
 #specifying the package makes it clearer where the function "lives",
 #but it adds typing. You can do it either way.
-rawdata <- readxl::read_excel(data_location)
+rawdata <- read.csv(data_location)
 
 #take a look at the data
 dplyr::glimpse(rawdata)
 
-#dataset is so small, we can print it to the screen.
-#that is often not possible.
-print(rawdata)
+head(rawdata)
+tail(rawdata)
+summary(rawdata)
 
-# looks like we have measurements for height (in centimeters) and weight (in kilogram)
+processeddata <- rawdata %>%
+  select(EXP, PCR, Method, Type, Dilution.Factor,FIN.Vol, Target, Ct, RH, Temp)%>% #select specific cols that may be interesting to compare
+  mutate(Target=as.factor(Target), Type=as.factor(Type), Method=as.factor(Method), EXP=as.factor(EXP)) %>% #mutate some cols to become specific data types
+  rename(Input=FIN.Vol) #Rename column for easier identification
+  
+##split from here and make 
+#Create supplementary dataframe due to all the Undetermined Ct Values from the reports.
+Undetermined<- processeddata%>%
+  filter(Ct=="Undetermined")
+#Save this too
+save_data_location <- here::here("data","processed_data","UndeterminedData.rds")
+saveRDS(Undetermined, file = save_data_location)
 
-# there are some problems with the data: 
-# There is an entry which says "sixty" instead of a number. 
-# Does that mean it should be a numeric 60? It somehow doesn't make
-# sense since the weight is 60kg, which can't happen for a 60cm person (a baby)
-# Since we don't know how to fix this, we need to remove the person.
-# This "sixty" entry also turned all Height entries into characters instead of numeric.
-# We need to fix that too.
-# Then there is one person with a height of 6. 
-# that could be a typo, or someone mistakenly entered their height in feet.
[…]

### 2021-10-29 code/processing_code/P2.Data.Asks.RMD
10 basics push - Work in prog
Basic push to not lose any data before refining and removing excess parts.

MC

---
title: "Asks"
author: "MYC"
date: "10/27/2021"
output: html_document
---
Loading all the default settings and preliminary programs.
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) #for data processing and all dplyr related programs
library(here)      #to set paths
library(tidymodels)      # for the recipes package, along with the rest of tidymodels
library(skimr)           # for variable summaries

```
List of asks:

* A vs B, 
* CDC vs NTP
* NTP 300 vs NTP 1000

```{r}
data%>%
  group_by(Detector, Method, Tool, Input)%>%
  summarize(n=n(),
            Ct=mean(Ct))
```

Path to Processed Data and loading of cleaned data
```{r load data}
data_location <- here::here("data","processed_data","P2.Processed.rds")
data<- readRDS(data_location)
glimpse(data)
```

# A vs B

Tools are used to sample different concntrations of seeded virus.
A recovered from -1 inoculated coupon area.
B recovered from -3 inoculated coupon area.
```{r}
data%>% 
  filter(Tool%in% c("A", "B"))%>%
  count(Method)

data%>% 
  filter(Tool%in% c("A", "B"), Input!="6400")%>%
  ggplot(aes(x=Tool, y= Ct))+
  geom_boxplot(aes(color=Method))+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  facet_grid(Detector~Input)+
  scale_y_reverse()+
  labs(y="Ct Value",
       x=NULL,
       title="Comparing A and B tool recovery")

```
Tool A has an average Ct recovery higher than B. This makes sense as the surfaces A recovered from is 2 logs higher (-1) than that of B (-3). A log drop is ~3Ct values we can see from the graphic that it trends as predicted.
Suggests that

# CDC vs NTP

Compare the different extraction methods Serial Dilutions
```{r}
data%>%
  mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
  filter(Tool=="SD", Input=="300")%>%
  ggplot(aes(x=Dilution, y=Ct, color=Method))+
  scale_y_reverse()+
  geom_point()+
  geom_smooth(aes(fill=Method),method=lm, alpha=0.20, fullrange=TRUE)+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
#  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  facet_wrap(Detector~Tool)+
  labs(title = "Serial Dilution comparisons between CDC and NTP Methods, Input at 300.")

```
Serial dilution data between CDC and NTP.
Noting that all SD are in the confirmed realm of under 27 Ct Value
CDC method is better than NTP at -1 concentration of the SD.
the -2 shows little difference between method recovery Ct values.
NTP appears to remain better than CDC at -3 and -4 levels.
Accounting for the SE looks like both methods are the same until the -2 dilution.
Might be significant difference at the -4 dilution.

Compare between extraction methods at input 300
```{r}
data%>%
  mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
  filter(Input=="300")%>%
  ggplot(aes(x=Dilution, y=Ct, color=Tool))+
  scale_y_reverse()+
  geom_jitter()+
  geom_smooth(method=lm)+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  facet_wrap(Method~Detector)+
  labs(title="Method extraction comparisons with standard input")

```

Compare between extraction methods with additional inputs
```{r}
data%>%
  mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
  filter(Input!="CTRL")%>%
  ggplot(aes(x=Dilution, y=Ct, shape=Tool, color=Input))+
  scale_y_reverse()+
  geom_jitter()+
  geom_smooth(method=lm)+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  facet_wrap(Method~Detector)+
  labs(title="Method extraction comparisons with all inputs and tools")
```

# NTP Input variations

Serial Dilution
```{r}
data%>%
  mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
  filter(Method=="NTP", Tool=="SD", Input!="CTRL", Input!="6800")%>%
  ggplot(aes(x=Dilution, y=Ct, color=Input, shape=Input))+
  scale_y_reverse()+
  geom_jitter()+
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  facet_wrap(.~Detector)+
  labs(title="Comparing Input levels on the Serial Dilituions with for the NTP method.")
```
NTP SD is similar to CDC SD in that the Ct values remain under 27, the most "confirmed" presence of nCoV-2.
Interesting! Looks like 300 input is really the lowest and as the inputs increase they kind of converge together at the -3 dilution level. 
Interpretation:
better detection at high copy levels with lower input; 
better detection at low copy levels with more input;

If we could determine how many copies are in the -3 dilution we may be able to better estimate the capacity of the particles.

NTP
```{r}
data%>%
  mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
  filter(Method=="NTP", Input=="300"|Input=="1000")%>%
  ggplot(aes(x=Dilution, y=Ct, shape=Input, color=Tool))+
  scale_y_reverse()+
  geom_jitter()+
  geom_smooth(method=glm, fullrange=TRUE)+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  facet_wrap(.~Detector)+
  labs(title="NTP extraction comparisons with all at 300 and 1000 input levels and all tools")
```

NTP
```{r}
data%>%
  mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
  filter(Method=="NTP", Tool!="SD", Input>=6000, Input!="CTRL")%>%
  ggplot(aes(x=Dilution, y=Ct, color=Input))+
  scale_y_reverse()+
  geom_boxplot(aes(fill=Input))+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  facet_grid(Tool~Detector)+
  labs(title="NTP extraction comparisons Eluents input levels and all tools")
```
Negatives show no detection. Positive controls are at -1 and have smaller Ct value than the A tools which recovered -1 inoculations. Gold line represents the threshold of 27

```{r}
data%>%
  filter(Input>=6000, Input!="CTRL")
```

### 2021-10-29 code/processing_code/P2.Data.Exploration.rmd
10 basics push - Work in prog
Basic push to not lose any data before refining and removing excess parts.

MC

---
title: "P2.Data.Exploration"
author: "MYC"
date: "10/25/2021"
output: 
  html_document:
    theme: cosmo
editor_options: 
  chunk_output_type: console
---
Loading all the default settings and preliminary programs.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) #for data processing and all dplyr related programs
library(here)      #to set paths
library(tidymodels)      # for the recipes package, along with the rest of tidymodels
library(skimr)           # for variable summaries

```
Path to Processed Data and loading of cleaned data
```{r load data}
data_location <- here::here("data","processed_data","P2.Processed.rds")
data<- readRDS(data_location)
glimpse(data)
```
Summary tables
```{r All Summary}
#making a summary table of proportions of Ct
data%>%
  count(Tool)%>%
  mutate(Proportions = n/sum(n)) #sample/sum of sample

data%>%
  skim()
```
## The Undetected Ct Values
Data Split out data for undetected
```{r Undetected}
#Create df of only samples without any Ct values
Undetected<-data%>%
  filter(is.na(Ct)) #filter for all NA's in Ct value-- this will mean not detected or error

skim(Undetected)

Undetected%>%
  filter(!grepl("*.CTRL", Tool))%>% #!grepl= removes anything ending with .CTRL in the Tool column
  group_by(Detector, Method, Input, Dilution)%>%
  count(Tool)%>%
  arrange(desc(n))%>%
  ggplot(aes(y=n,))+
  geom_col(position="dodge2", aes(x=Method, fill=Dilution ))+
  facet_grid(.~Detector)+
  labs(title="Number of Tool samples that were undetected after sampling and extradtion")
```
From the current dataset we see that at an input of 300 from tool B have a chance of being undetected.
The sample with most detection errors is the automated method recovering at the -3 log of solution for both probes.

## Back to the Data as a whole
```{r}
data%>%
  count(Method, Tool, Dilution, Input, sort=TRUE)

summary(data)
```

```{r}
ToolSummary<-data%>%
  select(Method, Detector, Tool, Input, Dilution, Ct)%>%
  filter(Method!="PCR", Tool%in%c("A","B","SD"))%>% #Remove PCR controls
  group_by(Detector, Method, Tool, Input)%>%
  summarise(n=n())%>%
  arrange(n,Detector)
print(ToolSummary)
```

```{r}
data%>%
  filter(Tool%in%c("SD", "NEG.CTRL","POS.CTRL"))%>%
  ggplot(aes(Dilution, Ct))+
  geom_line()+
  facet_wrap(~Tool)+
  scale_y_reverse()
```

SD Histogram between NTP and CDC
```{r}
data%>%
  filter(Tool%in%c("SD", "NEG.CTRL","POS.CTRL"))%>%
  arrange(Tool%in%c("NEG.CTRL", "SD", "POS.CTRL"),
          Input%in%c("300","1000","*"))%>%
  ggplot(aes(Dilution, Ct))+
  geom_jitter(aes(color=Input))+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  facet_wrap(~Method)+
  scale_y_reverse() #reverse the Y axis for Ct graphics, as lower Ct values indicate presence
```

### 2021-10-29 products/Part_2/DataExploration.Rmd
10 basics push - Work in prog
Basic push to not lose any data before refining and removing excess parts.

MC

@@
 __Making the UndeterminedData__
+
 see above but added filter() for Ct=="Undetermined" to pull all data with Undetermined as a reported value
@@
 
-####Run summaries
+#### Run summaries
 
-####Exploritory Figures
+#### Exploritory Figures
 
@@
 ![resultfigure2](../../results/resultfigure2.png)
-Interesting-- why is there such a group difference in Surface Sampling experiements versus Spike
+Interesting-- why is there such a group difference in Surface Sampling experiments versus Spike
 Oh makes sense it's SD- the serial dilution. We expect the SD to be like the standard curve and the recovery should have loss.
@@
 ![SS300](../../results/SS300.png)
+
+##### Local Polynomial Regression Smoothing (loess) with polynomial formula
+
+![SS300-Smooth](../../results/SS300Smooth.png)
+
+##### Density Plots -Experimental view
+
+###### Desnsity Output
+
+Doesn't take into account for the replicates in Sample facet. SD facet seems more in line.
+![SS300-Density](../../results/SS300-Density.png)
+
+###### ECDF analysis - Look into the data this way to see if __HOW__ different these are
+
+This is an ECDF between the Phase 2 runs. This compares CDC ONLY data between the two PCR runs as we adjusted changes. Done so I can include the CDC data with little change in code.
+Appears to have little difference, likely insignifant- will need to run stats.
+![SS300-ECDF.PCR](../../results/SS300-ECDF.PCR.png)
+
+###### 
+Exploring the
+
+![SS300-ECDF.Type](../../results/SS300-ECDF.Type.png)
+
+![SS300-ECDF.Method](../../results/SS300-ECDF.Method.png)

### 2021-10-29 products/manuscript/Manuscript.Rmd
10 basics push - Work in prog
Basic push to not lose any data before refining and removing excess parts.

MC

@@
 
-The structure below is a possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed.
-
 # Summary/Abstract
@@
 
-## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+##  Background
+The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants.
+Environmental reservoirs are a reported, indirect transmission route for viral pathogens.
+Although many viruses have a low infectious dose, they can be difficult to detect due to the low bioburden in the environment.
+The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control.
+Culture-based methods can be labor and time intensive.
+The use and limitations of commercial “target capture” (TC) technologies are varied compared to molecular approaches utilizing automated extraction.
+The objective of this study is improve outbreak response through rapid pathogen detection by evaluating a TC technology’s ability to capture SC2 from complex matrices.
 
-### Surfaces Preparation  
-Stainless-steel coupons (10x20in) were triple washed with detergent, rinsed with RO water, sprayed with 70% Ethanol and left to air dry.
-The dry coupons were wrapped in craft paper and autoclaved on a solid cycle.
-Before use, the coupons are unwrapped while in a BSC
-
-### Inoculum Preparation
-A serial dilution of heat-inactivated SARS-CoV-2 (~10^6 copies/uL, ATCC Strain) and Viral Transport Media (CDC protocol) was created
-
-### Inoculation of Surfaces
-
-## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
-
-Data was self recorded over several experimental runs of an on going series. 
-The current set has 379 observations and 23 variables. Additional data will be added into this set as experiments are completed throughout the course.
-The data itself is a combination of observed processing variables and the resulting RT-PCR data.
-Copy data may eventually be included.
-It is currently calculated in an excel document, but may be reproduced in this project for best practice).
-A copy of the data has been placed in the data folder of this repository.
-
-## Questions/Hypotheses to be addressed
+### Questions/Hypotheses to be addressed
 
@@
 
-# Methods and Results
+## Methods and Results
 
-_In most research papers, results and methods are separate. You can combine them here if you find it easier. You are also welcome to structure things such that those are separate sections._
+Using established standard methods for environmental sampling with a self-contained polyurethane swab, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR
 
 ## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+Reports per each rRT-PCR run were generated and quality checked on their respective machines before transfer to a Teams repository.
+Summary data was selected from the reports and added to additional environmental data gathered on based on the extraction date of the sample.
+Completed run data is then collated into one file and red into R.
 
-## Data import and cleaning
-
-Data comes out of different reports. This report is added to a supplemental report gathered while the samples were processed and added into one giant compilation then read into R.
+### Data import and cleaning
 

### 2021-10-29 code/analysis_code/analysisscript.R

@@ -5,6 +5,7 @@
 #and saves the results to the results folder
 
 #load needed packages. make sure they are installed.
+library(dplyr) #for data processing
 library(ggplot2) #for plotting
 library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
@@ -131,34 +132,92 @@ p3<-DirectCompare%>%
 ###look at figure
 plot(p3)
 
-###save figure
+###save figure SS300
 figure_file = here("results","SS300.png")
 ggsave(filename = figure_file, plot=p3)
 
 
 #### Let's focus only on the Surface sampling 
-DirectCompare%>%
+p4<-DirectCompare%>%
   filter(EXP=="SurfaceSampling", 
          Input=="300")%>% 
   ggplot(aes(x=Dilution.Factor, y=Ct))+
   geom_jitter()+
   geom_smooth(method="loess")+ #added smooth regression line
-  facet_grid(.~Type)
+  facet_grid(Method~Type)
+plot(p4)
+###save figure SS300Smooth
+figure_file = here("results","SS300Smooth.png")
+ggsave(filename = figure_file, plot=p4)
 
              
-
 #### Try Density plots
-mydata%>%
+DraftP5<-mydata%>%
   filter(EXP=="SurfaceSampling", 
          Input=="300")%>% 
   ggplot(aes(x=Ct, color=Method))+
   geom_density()+
   facet_grid(.~Type)
+plot(DraftP5)
+
+###save figure
+figure_file = here("results","SS300-Density.png")
+ggsave(filename = figure_file, plot=DraftP5)
+
 
 #### Try stat_ecdf()
-mydata%>%
+###Empirical Distribution Function
+### Concerned with observations rather than theory. Use data to create cumulative distribution
+### The more "separate" the lines are the more different they are from each other 
+DraftP6<-mydata%>%
   filter(EXP=="SurfaceSampling", 
-         Input=="300")%>% 
+         Input=="300",
+         Dilution.Factor!="NA")%>% 
   ggplot(aes(x=Ct, color=Method))+
   stat_ecdf()+
   facet_grid(Dilution.Factor~Type)
+
+###save figure
+figure_file = here("results","SS300-ECDF.Method.png")
+ggsave(filename = figure_file, plot=DraftP6)
+
+## no obvious difference between the methods within sample type
+
+DraftP7<-mydata%>%
+  filter(EXP=="SurfaceSampling", 
+         Input=="300",
[…]

### 2021-10-29 data/processed_data/Phase2ProcessingScript.R

@@ -1,0 +1,34 @@
+###############################
+# Phase 2 Processing script
+#
+#This script loads the raw phase 2 data as of 2021-10-22 data for processing and
+#cleaning it and saves it as Rds file in the processed_data folder
+
+#load needed packages.
+library(tidyverse) #for data processing
+library(here) #to set paths
+library(stringr)
+library(lubridate)
+#Confirm location using "here()" package for relative paths:
+here()
+#Establish data location, relative to top-level directory
+data_location <- here::here("data","raw_data","2021-10-23_Phase2.Update.csv")
+
+#load data. 
+RawP2 <- read.csv(data_location)
+
+glimpse(RawP2)
+#Data Shape up
+
+raw<-RawP2%>%
+  select(Phase, PCR, RH, TEMP, Plate, Method, Well, Tool, Input, Dilution, Sample.Name, Detector, Ct)%>%
+  mutate(PCR=as.Date(mdy(PCR)), Method=as.factor(Method), Ct=as.numeric(Ct), Detector=as.factor(Detector))%>%
+  filter(Tool!="")
+
+glimpse(raw)
+
+# save data as RDS
+# location to save file
+save_data_location <- here::here("data","processed_data","P2.Processed.rds")
+saveRDS(raw, file = save_data_location)
+

### 2021-10-30 code/processing_code/Cleaning_Reports.rmd
10 Clean up
Cleaning up and adding notations, making data connections more reproducible.

---
title: "Cleaning_Reports"
author: "MYC"
date: "10/29/2021"
output: html_document
---
Trying to determine the best way to read in excel reports for more automated data building.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(lubridate)
library(broom)
```
## Read in Reports 
Reports are generated in .csv outputs.
Would like to be able to read in each individual file from data\raw_data\Reports
```{r Read Reports}

```

## Scrape Data from reports
From read in data frames need to pull out a selection of data from A21:F12 (includes headers)

```{r Scraping Data}

```

### 2021-10-30 code/processing_code/ProcessingScript.Rmd
10 Clean up
Cleaning up and adding notations, making data connections more reproducible.

---
title: "Data Processing and Saving"
author: "MYC"
date: "10/29/2021"
output: 
  html_document:
    theme: cosmo
editor_options: 
  chunk_output_type:
---
Loading all the default settings and preliminary programs.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) #for data processing and all dplyr related programs
library(here)      #to set paths
library(lubridate) # for date manipulation
library(broom)    #for tidying up if needed

```
####
__Eventually this will read in the combined dataframe made from separate reports__

For now, this will import a self-maintained csv file to do initial analysis.

###

## Loading File
```{r Load All Data}
data_location <- here::here("data","raw_data","2021-10-23_All.Update.csv")
rawdata <- read.csv(data_location)
```
## Clean Data

### View Data
```{r view data}
dplyr::glimpse(rawdata)

head(rawdata)
tail(rawdata)
summary(rawdata)
```
## Wrangle Data
```{r wrangle data}
All.PCR<-rawdata%>%
  #select data cols that will be relevant for analysis
  select(Phase:Ct)%>%
  #change some of the data types into other data types
  mutate(Phase=as.factor(Phase), 
         Plate=as.factor(Plate),
         Method=as.factor(Method),
         Tool=as.factor(Tool),
         Input=as.numeric(Input),
         Dilution=as.numeric(Dilution),
         Detector=as.factor(Detector),
         Ct=as.numeric(Ct))%>%
  glimpse()
#leaving Ct as Character, for not
summary(All.PCR)
```
## Split Data and Save

### Phase 1 select Data (P1)
```{r P1 Select}
Phase1Data<-All.PCR%>%
  filter(Phase=="1")

summary(Phase1Data)
```
```{r P1 Save RDS}
save_data_location <- here::here("data","processed_data","P1-20211029.rds")

saveRDS(Phase1Data, file = save_data_location)
```

### Phase 2 select Data (P2)
```{r P2 Select}
Phase2Data<-All.PCR%>%
  filter(Phase=="2")

summary(Phase2Data)

```
```{r P2 Save RDS}

save_data_location <- here::here("data","processed_data","P2-20211029.rds")

saveRDS(Phase2Data, file = save_data_location)
```

Confirm Path to Processed Data and loading of cleaned data
```{r load data}
data_location <- here::here("data","processed_data",".rds")
data<- readRDS(data_location)
glimpse(data)
```

### 2021-10-30 products/Project_Parts/Part_1/Proposal.Rmd
10 Clean up
Cleaning up and adding notations, making data connections more reproducible.

---
title: "MADA Data Analysis Project Proposal"
subtitle: ""
author: Monica Chan
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: ../references.bib
csl: ../apa.csl
---

Applying MADA learned techniques to a RT-PCR dataset to determine the effectiveness of sample enhancement strategies for environmental surface sampling of low bio-burden viruses.

# Data Description
_What the data is_

Compilation of RT-PCR results for a couple of probes (N1 & N2) to detect recovered heat-inactivated SARS-CoV-2 samples that have undergone sample enhancement procedures.

_How was it collected_

Surface sampling, sample processing, RT-PCR run results, and result compilation were collected and performed by myself.

_How many observations_

The current set has 379 observations and 23 variables. Additional data will be added into this set as experiments are completed throughout the course.

_Additional information?_

Additional variables (Environmental and introduced) will be tested for effects.

# Data
```{r}
library(readr)
Data<-read_csv("../../data/raw_data/2021.08.18_Spike.csv")

str(Data)
summary(Data)
knitr::kable(Data[1:5,6:22] ,caption = 'First 5 Rows of data and a selection of variables table.')

```
## Questions to answer

There are 3 major aims of this:

1. Do the sample enhancement strategies work?
2. How well do they work compared to each other?
3. Are there any environmental variables recorded that have a positive impact on detection?

Testing several sample enhancement strategies (Nanoparticles and Bait & Capture) for molecular detection of heat-inactivated SARS-CoV-2 from stainless-steel surfaces. The primary aim of this work is to determine if any of the enhancement strategies increase the concentration of RNA copies in a sample that will express itself as a lower Ct value from the Rt-PCR output. Lower Ct values equate to more copies being present in the sample meaning less time needed to confirm presence.

In the case the sample enhancements work, we would then compare how well they worked against an automated extraction method. The degree of "wellness" will be determined by how low the Ct value results are produced after a sample has going through a specific method.

In addition to the comparison of enhancement strategies the effects of the tool used for sampling (macrofoam or knit-fiber), time held, and other environmental variables will be explored any significant effects, with interest in positive impacts to detection.

### Proposed Analysis

1. General data wrangling and cleaning
2. Summarize data for overview.
3. Compare Standard Curve Data for run differences
4. Compare Ct values between Methods, Tools, and Inputs (boxplots)
5. Adjust for difference if there were significant changed (ln?) between Serial dilutions on each PCR run.
6. Investigate any effects on the Ct recovery
7. Investigate loss between spiking experiment and sampling experiments

### Background Basics
Outbreak surveillance is contingent on the samples recovered from the environment and then identified. The introduction of molecular methods has helped in the rapid identification of species causing outbreaks, but these methods are only as good as the sample collected and used.

To insure a good sample can be taken, culture methods are generally required to grow all things present in the sample. Culture based methods are the gold standard in identification, but require resources and time to culture and a multitude of other factors may cause targeted organisms to die off or out competed during culturing.

Enter sample enhancement and a direct to molecular method workflow for quick ready to sequence samples.

Low-quality samples often are confounded by contamination or just low presence. Here I plan to utilize two types of sample enhancement strategies to see if these technologies work and how well they work against an automated sample extractor (Promega Maxwell 48). The first to be tested are Nanotrap Particles (Ceres Nanoscience), a hydrogel particle enhanced with specific affinity dyes that attract a target organism. The second strategy to be tested will be capture Baits (Arbor Sciences)- customized complementary baits for a target that can be used to select specific targets.

Though this project was created with Antimicrobial resistant organisms in mind, the 2020 pandemic halted research on non mission critical (Covid related) projects. Explaining the use of the SARS-CoV-2 virus in this study.

### 2021-11-02 code/analysis_code/01-ALLData.Rmd
clarify

---
title: "Analysis 01 - All Data"
author: "MYC"
date: "11/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) #for data processing
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving

```
Path to data
```{r Data}
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","ALL.Data-20211029.rds")
#read and load data. 
ALL<- readRDS(data_location)
```

# Summarize the Data

```{r}
mysummary <- summary(ALL)
```

```{r}
#re-shape of a data frame (for easier saving/showing in manuscript)
summary_df <- data.frame(do.call(cbind, lapply(ALL, summary)))
```
```{r}

```

```{r}
#save data frame table to file for later use in manuscript
summarytable_file <- here::here("products","Project_Parts","Part_3", "All.Summary.Table.rds")
saveRDS(summary_df, file = summarytable_file)
```

Make a scatterplot of data

```{r}
AllP1<-ALL %>% 
  filter(Method!="PCR", Tool=="SD"|Tool=="NEG.CTRL")%>%
  ggplot(aes(x=Dilution, y=Ct)) + 
  geom_jitter(aes(color=Method)) + 
  #we also add a glm line to it- using glm due to the continuous nature of the dilutions??
  geom_smooth(method='glm',alpha=0.2, aes(fill=Method, color=Method))+
  facet_grid(Detector~.)+
  scale_y_reverse()+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  labs(title="Serial Dilution Comparisons within ALL the data")

plot(AllP1)
```

```{r}
#save figure
figure_file = here("results","ALL-SD-Plot.png")
ggsave(filename = figure_file, plot=AllP1) 
```

```{r ALL Data fit and stat analysis}

```
######################################
#Data fitting/statistical analysis
######################################
```{r}
# fit glm.
# Ct~X= odds of Ct as response variable
ALL.glm.fit<-glm(Ct~Dilution+Detector+Method, data=ALL)
#look at fit
print(tidy(ALL.glm.fit))

step(ALL.glm.fit, test="LRT")
```
```{r}
# save fit results table  
table_file = here("results", "resulttable.rds")
saveRDS(lmtable, file = table_file)

```

### 2021-11-02 code/analysis_code/02-P1.Data.rmd
clarify

---
title: "Analysis 02 - Phase 1 Data"
author: "MYC"
date: "11/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) #for data processing
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving

```
Path to data
```{r Data}
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","P1-20211029.rds")
#read and load data. 
P1<- readRDS(data_location)
```

# Summarize the Data

```{r}
P1mysummary <- summary(P1)
```

```{r}
#re-shape of a data frame (for easier saving/showing in manuscript)
P1summary_df <- data.frame(do.call(cbind, lapply(P1, summary)))
```
```{r}

```

```{r}
#save data frame table to file for later use in manuscript
summarytable_file <- here::here("products","Project_Parts","Part_3", "P1.Summary.Table.rds")
saveRDS(P1summary_df, file = summarytable_file)
```

Make a scatter plot of data
```{r}
P1P1<-P1 %>% 
  filter(Method!="PCR", Tool=="SD"|Tool=="NEG.CTRL")%>%
  ggplot(aes(x=Dilution, y=Ct)) + 
  geom_jitter(aes(color=Method)) + 
  #we also add a linear regression line to it
  geom_smooth(method='glm',alpha=0.2, aes(fill=Method, color=Method))+
  facet_grid(Detector~.)+
  scale_y_reverse()+geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  labs(title="Serial Dilution Comparisons within Phase 1 data")

print(P1P1)
```
```{r}
#save figure
figure_file = here("results","P1-SD-Plot.png")
ggsave(filename = figure_file, plot=P1P1) 
```

```{r ALL Data fit and stat analysis}

```
######################################
#Data fitting/statistical analysis
######################################
```{r}
# fit glm.
# Ct~X= odds of Ct as response variable
P1.glm.fit<-glm(Ct~Dilution+Detector+Method, data=P1)
#look at fit
print(tidy(P1.glm.fit))

step(P1.glm.fit, test="LRT")
```
```{r}
# save fit results table  
table_file = here("results", "resulttable.rds")
saveRDS(lmtable, file = table_file)

```

### 2021-11-02 code/analysis_code/03-P2.Data.rmd
clarify

---
title: "Analysis 03 - Phase 2 Data"
author: "MYC"
date: "11/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) #for data processing
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving

```
Path to data
```{r Data}
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","P2-20211029.rds")
#read and load data. 
P2<- readRDS(data_location)
```

# Summarize the Data

```{r}
p2mysummary <- summary(P2)
```

```{r}
#re-shape of a data frame (for easier saving/showing in manuscript)
P2summary_df <- data.frame(do.call(cbind, lapply(P2, summary)))
```
```{r}

```

```{r}
#save data frame table to file for later use in manuscript
summarytable_file <- here::here("products","Project_Parts","Part_3", "P2.Summary.Table.rds")
saveRDS(P2summary_df, file = summarytable_file)
```

Make a scatterplot of data

```{r}
P2P1<-P2 %>% 
  filter(Method!="PCR", Tool=="SD"|Tool=="NEG.CTRL")%>%
  ggplot(aes(x=Dilution, y=Ct)) + 
  geom_jitter(aes(color=Method)) + 
  #we also add a linear regression line to it
  geom_smooth(method='glm',alpha=0.2, aes(fill=Method, color=Method))+
  facet_grid(Detector~.)+
  scale_y_reverse()+
  geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
  labs(title="Serial Dilution Comparisons within Phase 1 data")

print(P2P1)
```
```{r}
#save figure
figure_file = here("results","P2-SD-Plot.png")
ggsave(filename = figure_file, plot=P2P1) 
```

```{r ALL Data fit and stat analysis}

```
######################################
#Data fitting/statistical analysis
######################################
```{r}
# fit glm.
# Ct~X= odds of Ct as response variable
P2.glm.fit<-glm(Ct~Dilution+Detector, data=P2P1)
#look at fit
print(tidy(P2.glm.fit))

step(P2.glm.fit, test="LRT")
```

### 2021-11-02 code/processing_code/ProcessingScript.Rmd
clarify

@@
 ####
-__Eventually this will read in the combined dataframe made from separate reports__
+__Eventually this will read in the combined data frame made from separate reports__
 
@@
 ```
+```{r All Data}
+save_data_location <- here::here("data","processed_data","ALL.Data-20211029.rds")
+
+saveRDS(All.PCR, file = save_data_location)
+```
+
 ## Split Data and Save

### 2021-11-02 products/Project_Parts/Part_3/Part3.Rmd
clarify

--- 2021-10-29 products/Part_2/DataExploration.Rmd
@@
 ---
-title: "Data Exploration"
+title: "Part 3"
 subtitle: ""

### 2021-11-02 products/manuscript/Manuscript.Rmd
clarify

@@
 
-There are 3 major aims of this:
+There are several major aims of this:
 
@@
 3. Are there any environmental variables recorded that have a positive impact on detection?
+4. Can this data be modeled to back calculate a sample's original concentration based on the Ct values recovered?
 
@@
 
-Using established standard methods for environmental sampling with a self-contained polyurethane swab, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR
+### Experiment setup and conditions
+
+Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR
 
@@
 Summary data was selected from the reports and added to additional environmental data gathered on based on the extraction date of the sample.
-Completed run data is then collated into one file and red into R.
+
+Completed run data is then collated into one file and read into R./A composited data set was compiled and saved as a csv and used as the base data for the project
 
@@
 
-processingscript.R was modified to do initial cleaning and preparation of the dataset for further exploration.Data is also saved.
+processingscript.R was modified to do initial cleaning and preparation of the dataset for further exploration.
 
-## Exploratory analysis
+Data was imported and cleaned.
+Data was then split between the two different experimental types (P1 and P2).
+
+This can be found in processing_code/ProcessingScript.RMD
+
+### Data exploration
 
@@
 Depending on the total number of variables in your dataset, explore all or some of the others.
-FIgures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data._
 
@@
 Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+
+#### All Data
+
+Overview - split between part 1 and part 2
+
+```{r ALLsummarytable,  echo=FALSE}
+resulttable=readRDS("../../results/ALLsummarytable.rds")
+knitr::kable(ALLresulttable, caption = 'ALL Data summary table.')
+```
+
+##### Serial Dilutions confirmation
+
+Summary table 
+scatter plot of SDs based ond dilution factors
+
+#### Phase 1
+
+### Phase 2
+
+### Data Analysis
+
+### Data Modeling
+
+## Exploratory analysis
 

### 2021-11-04 ARCHIVED/Manuscript.Rmd
Update
Tried to really archive template items
Added clarification to read me
saved figures and added to sections in report.

--- 2021-11-02 products/manuscript/Manuscript.Rmd
@@
 # Summary/Abstract
-_Write a summary of your project._  
+
 Applying MADA learned techniques to a RT-PCR dataset to determine the effectiveness of sample enhancement strategies for environmental surface sampling of low bio-burden viruses.
@@
 
-```{r ALLsummarytable,  echo=FALSE}
-resulttable=readRDS("../../results/ALLsummarytable.rds")
-knitr::kable(ALLresulttable, caption = 'ALL Data summary table.')
-```
-
 ##### Serial Dilutions confirmation
@@
 scatter plot of SDs based ond dilution factors
-
+```{r}
+knitr::include_graphics("../../results/ALL-SD-Plot.png")
+```
 #### Phase 1
 
+```{r}
+knitr::include_graphics("../../results/P1-SD-Plot.png")
+```
+
 ### Phase 2
+
+```{r}
+knitr::include_graphics("../../results/P2-SD-Plot.png")
+```
 
@@
 
-```{r summarytable,  echo=FALSE}
-resulttable=readRDS("../../results/summarytable.rds")
-knitr::kable(resulttable, caption = 'Data summary table.')
-```
-
 Figure \@ref(fig:resultfigure) shows a scatterplot figure produced by one of the R scripts.
 
-```{r resultfigure,  fig.cap='Analysis figure.', echo=FALSE}
+```{r resultfigure, eval=FALSE, fig.cap='Analysis figure.', include=FALSE}
 knitr::include_graphics("../../results/resultfigure.png")
@@
 Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.
-
-```{r resulttable,  echo=FALSE}
-resulttable=readRDS("../../results/resulttable.rds")
-knitr::kable(resulttable, caption = 'Linear model fit table.')
-```
 

### 2021-11-04 code/analysis_code/01-ALLData.Rmd
Update
Tried to really archive template items
Added clarification to read me
saved figures and added to sections in report.

@@
 ```{r}
-
+ALL%>%
+  filter(Method!="PCR", Dilution!="NA")%>%
+  group_by(Detector, Method, Dilution, Input)%>%
+  summarise(n=n(),
+            "Average Ct"= mean(Ct)
+            )
 ```
@@
 # save fit results table  
-table_file = here("results", "resulttable.rds")
-saveRDS(lmtable, file = table_file)
+table_file = here("results", "AllGLM.resulttable.rds")
+saveRDS(ALL.glm.fit, file = table_file)
 

### 2021-11-04 code/analysis_code/03-P2.Data.rmd
Update
Tried to really archive template items
Added clarification to read me
saved figures and added to sections in report.

@@
   geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
-  labs(title="Serial Dilution Comparisons within Phase 1 data")
+  labs(title="Serial Dilution Comparisons within Phase 2 data")
 

### 2021-11-04 code/processing_code/P2.Data.Asks.RMD
10 Clean up
Cleaning up and adding notations, making data connections more reproducible.
Update
Tried to really archive template items
Added clarification to read me
saved figures and added to sections in report.

@@
 
+Path to Processed Data and loading of cleaned data
+```{r load data}
+data_location <- here::here("data","processed_data","P2-20211029.rds")
+data<- readRDS(data_location)
+glimpse(data)
+```
 ```{r}
@@
   group_by(Detector, Method, Tool, Input)%>%
+  filter(Method!="PCR")%>%
   summarize(n=n(),
-            Ct=mean(Ct))
+            Ct=mean(Ct))%>%
+  mutate(tools=(n/2))
 ```
-
-Path to Processed Data and loading of cleaned data
-```{r load data}
-data_location <- here::here("data","processed_data","P2.Processed.rds")
-data<- readRDS(data_location)
-glimpse(data)
-```
-
 # A vs B
@@
 
-data%>% 
+AsksP2P1<-data%>% 
   filter(Tool%in% c("A", "B"), Input!="6400")%>%
@@
        x=NULL,
-       title="Comparing A and B tool recovery")
+       title="Phase 2 Comparing A and B tool recovery")
 
+print(AsksP2P1)
+
+```
+```{r}
+#save figure
+figure_file = here("results","ASK-P2P1-Tools.png")
+ggsave(filename = figure_file, plot=AsksP2P1) 
 ```
@@
 ```{r}
-data%>%
+AsksP2P2<-data%>%
   mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
@@
 
+print(AsksP2P2)
 ```
+```{r}
+#save figure
+figure_file = here("results","ASK-P2P2-Methods.png")
+ggsave(filename = figure_file, plot=AsksP2P2) 
+```
+
 Serial dilution data between CDC and NTP.
@@
 ```{r}
-data%>%
+AsksP2P3<- data%>%
   mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
@@
   facet_wrap(Method~Detector)+
-  labs(title="Method extraction comparisons with standard input")
+  labs(title="Method extraction comparisons with standard input (300)")
 
+print(AsksP2P3)
+```
+```{r}
+#save figure
+figure_file = here("results","ASK-P2P3-Methods0300.png")
+ggsave(filename = figure_file, plot=AsksP2P3) 
 ```
@@
 ```{r}
-data%>%
+AsksP2P4<-data%>%
   mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
@@
   labs(title="Method extraction comparisons with all inputs and tools")
+
+print(AsksP2P4)
+```
+```{r}
+#save figure
+figure_file = here("results","ASK-P2P4-MethodsALL.png")
+ggsave(filename = figure_file, plot=AsksP2P4) 
 ```
@@
 ```{r}
-data%>%
+AsksP2P5<-data%>%
   mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
   filter(Method=="NTP", Tool=="SD", Input!="CTRL", Input!="6800")%>%
-  ggplot(aes(x=Dilution, y=Ct, color=Input, shape=Input))+
+  ggplot(aes(x=Dilution, y=Ct, color=Input))+
   scale_y_reverse()+
@@
   labs(title="Comparing Input levels on the Serial Dilituions with for the NTP method.")
+
+print(AsksP2P5)
 ```
+```{r}
+#save figure
+figure_file = here("results","ASK-P2P5-NTP-SD.png")
+ggsave(filename = figure_file, plot=AsksP2P5) 
+```
+
 NTP SD is similar to CDC SD in that the Ct values remain under 27, the most "confirmed" presence of nCoV-2.
@@
 ```{r}
-data%>%
+AsksP2P6<-data%>%
   mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
   filter(Method=="NTP", Input=="300"|Input=="1000")%>%
-  ggplot(aes(x=Dilution, y=Ct, shape=Input, color=Tool))+
+  ggplot(aes(x=Dilution, y=Ct, color=Tool))+
   scale_y_reverse()+
@@
   labs(title="NTP extraction comparisons with all at 300 and 1000 input levels and all tools")
+print(AsksP2P6)
+```
+```{r}
+#save figure
+figure_file = here("results","ASK-P2P6-NTP.png")
+ggsave(filename = figure_file, plot=AsksP2P6) 
 ```
@@
 NTP
+NTP extraciton comparisons wluent input levels and all tools
 ```{r}
@@
   filter(Method=="NTP", Tool!="SD", Input>=6000, Input!="CTRL")%>%
-  ggplot(aes(x=Dilution, y=Ct, color=Input))+
+  ggplot(aes(x=Dilution, y=Ct))+
   scale_y_reverse()+
-  geom_boxplot(aes(fill=Input))+
+  geom_boxplot(aes())+
   geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold

### 2021-11-12 code/processing_code/ProcessingScript.Rmd
Clean up and Temp Code
Clean up of unused scripts and removal of knitted items. Temp Code added-- working on doing a completely automated data import from reports.

@@
 ---
+This script loads the 2021-10-23_All.Update.csv, a composite of reports, from `raw-data`.
+The script further processes, cleans, splits, and saves the data as rds files in the `processed_data` folder.
+
 Loading all the default settings and preliminary programs.

### 2021-11-12 code/processing_code/ProcessingScript2.rmd
Clean up and Temp Code
Clean up of unused scripts and removal of knitted items. Temp Code added-- working on doing a completely automated data import from reports.

---
title: "ProcessingScript2"
author: "MYC"
date: "11/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr) #for data processing
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(reshape2)

```
Path to data
```{r Data}
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","P2-20211029.rds")
#read and load data. 
P2<- readRDS(data_location)
```

```{r}
summary(P2)

melt(P2, 
     id.vars = c("Detector", "Method", "Tool"),
     measure.vars = c("Input","Ct"))%>%
  summarise(n=n())

```
```{r}
dcast(P2, Method+Tool+Input+Dilution+Ct~Tool)%>%
  ggplot(aes(y=Ct, x=Dilution))+
  geom_jitter(aes(color=Method))
```

### 2021-11-12 code/processing_code/TEMP/Cleaning_Reports.rmd
Clean up and Temp Code
Clean up of unused scripts and removal of knitted items. Temp Code added-- working on doing a completely automated data import from reports.

---
title: "Cleaning_Reports"
author: "MYC"
date: "10/29/2021"
output: html_document
---
Trying to determine the best way to read in excel reports for more automated data building.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(lubridate)
library(fs)
```
## Read in Reports 
Reports are generated in .csv outputs.
Would like to be able to read in each individual file from data\raw_data\Reports
```{r Path to Reports}
#set the directory/file location of reports
data_dir<-(here::here("data","raw_data","Reports"))

#create path for all csv files in directory of reports
csv_files<-fs::dir_ls(data_dir, # shows list of files in directory
                      regexp = "\\.csv$") #select only.csv files 
```

## Read section of reports
From read in data frames need to pull out a selection of data from A21:F12 (includes headers)

```{r Read and select from Reports}

# read a single report in as a data frame
Report01<-read.csv(csv_files[1],nrow=94, skip=28)[,1:6] #range finding phases
Report02<-read.csv(csv_files[2],nrow=94, skip=28)[,1:6]#range finding phases
Report03<-read.csv(csv_files[3],nrow=94, skip=28)[,1:6]#range finding phases
Report04<-read.csv(csv_files[4],nrow=94, skip=28)[,1:6]#range finding phases
Report05<-read.csv(csv_files[5],nrow=94, skip=28)[,1:6]#range finding phases
Report06<-read.csv(csv_files[6],nrow=76, skip=28)[,1:6] #starts Phase 2
Report07<-read.csv(csv_files[7],nrow=76, skip=28)[,1:6] #starts Phase 2
Report08<-read.csv(csv_files[8],nrow=76, skip=28)[,1:6] #starts Phase 2
Report09<-read.csv(csv_files[9],nrow=76, skip=28)[,1:6] #starts Phase 2
Report10<-read.csv(csv_files[10],nrow=76, skip=28)[,1:6] #starts Phase 2
Report11<-read.csv(csv_files[11],nrow=76, skip=28)[,1:6] #starts Phase 2

```

```{r All Data}
#merge all same variables from all reports
AllData<-rbind(Report01,
      Report02,
      Report03,
      Report04,
      Report05,
      Report06,
      Report07,
      Report08,
      Report09,
      Report10,
      Report11
      )
```

```{r P1}
#selective merge of data for part 1
P1<-rbind(Report01,
      Report02,
      Report03,
      Report04,
      Report05
      )
```

```{r P2}
#selective merge of data part 2
P2<-rbind(
      Report06,
      Report07,
      Report08,
      Report09,
      Report10,
      Report11
      )
```

### 2021-11-12 code/processing_code/TEMP/P2.Data.Exploration.rmd
Clean up and Temp Code
Clean up of unused scripts and removal of knitted items. Temp Code added-- working on doing a completely automated data import from reports.

--- 2021-10-29 code/processing_code/P2.Data.Exploration.rmd
@@
     theme: cosmo
-editor_options: 
-  chunk_output_type: console
 ---

### 2021-11-13 code/analysis_code/03-P2.Data.rmd
Update References
Select and add References to manuscript

@@
 summarytable_file <- here::here("products","Project_Parts","Part_3", "P2.Summary.Table.rds")
-saveRDS(P2summary_df, file = summarytable_file)
+#saveRDS(P2summary_df, file = summarytable_file)
 ```

### 2021-11-13 code/analysis_code/04-Mods.Rmd
Update References
Select and add References to manuscript

---
title: "Exploritory Modeling and Effects- P2 only"
author: "MYC"
date: "11/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) #for data processing
library(here) #to set paths
library(tidymodels)      # for the recipes package, along with the rest of tidymodels
library(skimr)           # for variable summaries
library(rpart)
library(tune)
library(yardstick)
```
Path to data
```{r Data}
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","P2-20211029.rds")
#read and load data. 
P2<- readRDS(data_location)

```
View proportions of the samples with Method
```{r}
P2%>%
  count(Method)%>%
  mutate(Proportions =n/sum(n))
```

View proportions of the samples within NTP method
```{r}
P2%>%
  filter(Method=="NTP")%>%
  count(Input)%>%
  mutate(Proportions =n/sum(n))
```
Data Splitting
Split into traiining and testing set
```{r}
# Fix the random numbers by setting the seed 
# This enables the analysis to be reproducible when random numbers are used 
set.seed(123)
# Put 3/4 of the data into the training set; this leaves 1/4 of the data to be used to test 
data_split <- initial_split(P2, prop = 3/4)
## Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```
Create CV object
```{r}
CV_train<- 
  vfold_cv(train_data, v=5, repeats= 5, strata = "Method")
```
Create a recipe
```{r}
fit_rec<-
  recipe(Method~., data = train_data)%>%
  step_dummy(all_nominal())
```
Null Model ? - do i need this?
```{r}
#Ct is a continuous outcome-- need to do RMSE as performance metric

RMSE_null_train<-sqrt(sum((train_data$Ct - mean(train_data$Ct))^2)/nrow(train_data))
RMSE_null_test<-sqrt(sum((test_data$Ct - mean(test_data$Ct))^2)/nrow(test_data))

print(RMSE_null_train)
print(RMSE_null_test)
```
Tree model setup
```{r}
tree_model<-decision_tree()%>%
  set_args(cost_complexity(tune(), tree_depth(tune(), min_n=tune())))%>%
  set_engine("rpart")%>%
  set_mode("regression")
```
```{r}
#set wf
tree_wf<- workflow()%>%
  add_model(tree_model)%>%
  add_recipe(fit_rec)
```
Tree Tuning
```{r}
#tuning grid
tree_grid <- dials::grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)
#tune the model
tree_tune_res <- tree_wf %>% 
  tune::tune_grid(
    resamples = CV_train,
    grid = tree_grid,
    metrics = yardstick::metric_set(rmse) 
  )

```

Try LASSO
```{r}
#model
lasso_model <- linear_reg() %>%
  set_mode("regression") %>%           
  set_engine("glmnet") %>%
  set_args(penalty = tune(), mixture = 1) #mixture = 1 means we use the LASSO model

#workflow
lasso_wf <- workflow() %>%
  add_model(lasso_model) %>% 
  add_recipe(fit_rec)
```
LASSO Tune
```{r}
#tuning grid
lasso_reg_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30))

#tune model
lasso_tune_res <- lasso_wf %>% 
  tune_grid(resamples = CV_train,
            grid = lasso_reg_grid,
            control = control_grid(save_pred = TRUE),
            metrics = metric_set(rmse)
            )

```
```{r}
#see a plot of performance for different tuning parameters
lasso_tune_res %>% autoplot()
```

```{r}

Ct~Input*Method

lm_mod<-linear_reg()%>%
  set_engine("lm")

  lm_mod%>%
  fit(Ct~Input*Method, data=P2)%>%
    tidy()

```

### 2021-11-13 products/manuscript/Manuscript.Rmd
Update References
Select and add References to manuscript

@@
 output:
+  bookdown::html_document2: 
+    toc: false
   bookdown::word_document2: 
@@
     number_sections: true
-  bookdown::html_document2: 
-    toc: false
 bibliography: ../references.bib
@@
 The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants.
-Environmental reservoirs are a reported, indirect transmission route for viral pathogens.
+Environmental reservoirs are a reported, indirect transmission route for viral pathogens [@Marcenac2021-sz].
 Although many viruses have a low infectious dose, they can be difficult to detect due to the low bioburden in the environment.
-The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control.
+The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control [@Paton2021-jc].
 Culture-based methods can be labor and time intensive.
@@
 
-Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR
+Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR [@Wittwer2012-vt][@Shah2021-mb].
 
@@
 
-processingscript.R was modified to do initial cleaning and preparation of the dataset for further exploration.
+processing script.R was modified to do initial cleaning and preparation of the data set for further exploration.
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-You should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
+Direction was influenced by [@Weir2016-sj] on "Effect fo Surface Sampling and Recovery of Viruses and No-Spore-Forming Bacteria on a Quantitative Microbial Risk Assessment Model for Fomites".
 
-_Continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-Plots produced here can be scatterplots, boxplots, violinplots, etc.
-Tables can be simple 2x2 tables or larger ones._
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-t-tests, simple regression model with 1 predictor, etc.) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
+Atempt to do as [@Brandolini2021-ja]
 
@@
 Summary table 
-scatter plot of SDs based ond dilution factors
-
+scatter plot of SDs based on dilution factors
+```{r}
+knitr::include_graphics("../../results/ALL-SD-Plot.png")
+```
 #### Phase 1
 
+```{r}
+knitr::include_graphics("../../results/P1-SD-Plot.png")
+```
+
 ### Phase 2
+
+```{r}
+knitr::include_graphics("../../results/P2-SD-Plot.png")
+```
+#### Phase 2 Comparing A and B Tool Recovery 
+```{r}
+knitr::include_graphics("../../results/ASK-P2P1-Tools.png")
+```
+Tool A has an average Ct recovery higher than B.
+This makes sense as the surfaces A recovered from is 2 logs higher (-1) than that of B (-3).
+A log drop is ~3Ct values we can see from the graphic that it trends as predicted.
+Suggests that
+
+#### Phase 2 Comparing Method Serial Dilutions (CDC vs NTP) with standard input of 300uL
+```{r}
+knitr::include_graphics("../../results/ASK-P2P2-Methods.png")
+```
+Serial dilution data between CDC and NTP.
+Noting that all SD are in the confirmed realm of under 27 Ct Value CDC method is better than NTP at -1 concentration of the SD.
+the -2 shows little difference between method recovery Ct values.
+NTP appears to remain better than CDC at -3 and -4 levels.
+Accounting for the SE looks like both methods are the same until the -2 dilution.
+Might be significant difference at the -4 dilution.
+
+#### Phase 2 Comparing Methods with standard input of 300uL
+```{r}
+knitr::include_graphics("../../results/ASK-P2P3-Methods0300.png")
+```
+
+#### Phase 2 Method extraction comparisons with all inputs and tools
+```{r}
+knitr::include_graphics("../../results/ASK-P2P4-MethodsALL.png")
+```
+
+#### Phase 2 Nanoparticles only - Compare detectors with dilution and input in Serial Dilutions
+```{r}
+knitr::include_graphics("../../results/ASK-P2P5-NTP-SD.png")
+```
+
+#### Phase 2 Nanoparticles only - All tools at 300 and 1000ul input levels
+```{r}
+knitr::include_graphics("../../results/ASK-P2P6-NTP.png")
+```
 
@@
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
 Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.
-
-```{r resulttable,  echo=FALSE}
-resulttable=readRDS("../../results/resulttable.rds")
-knitr::kable(resulttable, caption = 'Linear model fit table.')
-```
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
+Took
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
+Need more replicates, but this is a time consuming processes just to get the data.
+
+Additional materials need to be considered [@Jansson2020-cg].
 
 ## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [@Leek2015a] discusses types of analyses.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.
 

### 2021-11-13 code/analysis_code/03-P2.Data.rmd
Fixed manuscript and processing scripts
removing unfinished chunks

@@
 p2mysummary <- summary(P2)
+
+print(p2mysummary)
 ```
@@
 P2summary_df <- data.frame(do.call(cbind, lapply(P2, summary)))
+
+print(P2summary_df)
 ```
 ```{r}
-
+P2%>%
+  filter(Dilution!="NA", Tool!="SD", Tool!="*.CTRL")%>%
+  mutate(
+    Ct.Cat = ifelse(Ct >=38,"Detected","Positive"),
+    Ct.Cat=factor(Ct.Cat))%>%
+  group_by(Dilution,Method,Ct.Cat)%>%
+  summarise(n=n(),
+            mean=mean(Ct))%>%
+  arrange(Dilution,mean)
 ```
@@
 ```
+
+```{r Table Best Ct comparing spike and surface recovery}
+#Summary of Tool A compared to -1 Serial Dilution Recovery
+P1.1Dilution.Summary<-P2%>%
+  filter(Dilution=="-1")%>%
+  mutate(
+    Ct.Cat = ifelse(Ct >=38,"Detected","Positive"),
+    Ct.Cat=factor(Ct.Cat))%>%
+  group_by(Tool,Method,Ct.Cat)%>%
+  summarise(n=n(),
+            mean=mean(Ct))%>%
+  arrange(mean)%>%
+  na.omit(C.Cat)
+```
+```{r}
+#save data frame table to file for later use in manuscript
+summarytable_file <- here::here("products","supplement", "P2.-1Dilution.Table.rds")
+saveRDS(P1.1Dilution.Summary, file = summarytable_file)
+```
+A difference of 2 Ct values when compared by methods for each tool type. CDC was better when directly extracted. NTP performed better during a spiked tool and surface recovery procedures. 
+Ct loss between Spike and SS recovery are about 3 Ct values, equivalent to a log loss of copies.
 
@@
 ######################################
-```{r}
-# fit glm.
-# Ct~X= odds of Ct as response variable
-P2.glm.fit<-glm(Ct~Dilution+Detector, data=P2P1)
-#look at fit
-print(tidy(P2.glm.fit))
-
-step(P2.glm.fit, test="LRT")
-```

### 2021-11-13 code/analysis_code/04-Mods.Rmd
Fixed manuscript and processing scripts
removing unfinished chunks

@@
 ```
+
+```{r}
+P2%>%
+  mutate(Norm.Ct=Ct/Input)
+```
+
 Data Splitting
-Split into traiining and testing set
+Split into training and testing set
 ```{r}
@@
 
-  lm_mod%>%
-  fit(Ct~Input*Method, data=P2)%>%
-    tidy()
+lm_fit<-  lm_mod%>%
+  fit(Ct~Input*Method, data=P2)
+lm_fit
+```
+```{r}
+new_points<- expand.grid(initial_volume=20,
+                         food_regime=)
+```
 
+```{r}
+  P2%>%
+  filter(Dilution!="NA", Tool!="SD", Tool!="*.CTRL")%>%
+  mutate(
+    Ct.Cat = ifelse(Ct >=38,"Detected","Positive"),
+    Ct.Cat=factor(Ct.Cat))%>%
+  group_by(Dilution,Method,Ct.Cat)%>%
+  summarise(n=n(),
+            mean=mean(Ct))%>%
+  arrange(Dilution,mean)
 ```
+```{r Table Best Ct comparing spike and surface recovery}
+  P2%>%
+  filter(Dilution=="-1")%>%
+  mutate(
+    Ct.Cat = ifelse(Ct >=38,"Detected","Positive"),
+    Ct.Cat=factor(Ct.Cat))%>%
+  group_by(Tool,Method,Ct.Cat)%>%
+  summarise(n=n(),
+            mean=mean(Ct))%>%
+  arrange(mean)
+```

### 2021-11-13 code/processing_code/ProcessingScript.Rmd
Fixed manuscript and processing scripts
removing unfinished chunks

@@
 date: "10/29/2021"
-output: 
-  html_document:
-    theme: cosmo
-editor_options: 
-  chunk_output_type:
+output: html_document
 ---
@@
 ```
-
-Confirm Path to Processed Data and loading of cleaned data
-```{r load data}
-data_location <- here::here("data","processed_data",".rds")
-data<- readRDS(data_location)
-glimpse(data)
-```

### 2021-11-13 products/manuscript/Manuscript.Rmd
Fixed manuscript and processing scripts
removing unfinished chunks

@@
 
-In addition to the comparison of enhancement strategies the effects of the tool used for sampling (macrofoam or knit-fiber), time held, and other environmental variables will be explored any significant effects, with interest in positive impacts to detection.
+In addition to the comparison of enhancement strategies the effects of the tool used for sampling (macrofoam or knit-fiber), time held, and other environmental variables will be explored any significant effects, with interest in positive impacts to detection [@Jansson2020-cg].
 

### 2021-11-14 code/analysis_code/01-ALLData.Rmd
Adjust and re-knit analysis script. Re-do modeling still working on it

--- 2021-11-02 code/analysis_code/01-ALLData.Rmd
@@
 mysummary <- summary(ALL)
+print(mysummary)
 ```
 
-```{r}
+```{r eval=FALSE, include=FALSE}
 #re-shape of a data frame (for easier saving/showing in manuscript)
@@
 ```{r}
-
+All.Summarize<-ALL%>%
+  filter(Method!="PCR", Dilution!="NA")%>%
+  group_by(Detector, Method, Dilution, Input)%>%
+  summarise(n=n(),
+            "Average Ct"= mean(Ct)
+            )
 ```
@@
 #save data frame table to file for later use in manuscript
-summarytable_file <- here::here("products","Project_Parts","Part_3", "All.Summary.Table.rds")
-saveRDS(summary_df, file = summarytable_file)
+summarytable_file <- here::here("products","supplement", "All.Summary.Table.rds")
+saveRDS(All.Summarize, file = summarytable_file)
 ```
@@
 ######################################
-```{r}
+```{r eval=FALSE, include=FALSE}
 # fit glm.
@@
 ```
-```{r}
+```{r eval=FALSE, include=FALSE}
 # save fit results table  
-table_file = here("results", "resulttable.rds")
-saveRDS(lmtable, file = table_file)
+table_file = here("results", "AllGLM.resulttable.rds")
+saveRDS(ALL.glm.fit, file = table_file)
 

### 2021-11-14 code/analysis_code/02-P1.Data.rmd
Adjust and re-knit analysis script. Re-do modeling still working on it

@@
 #save data frame table to file for later use in manuscript
-summarytable_file <- here::here("products","Project_Parts","Part_3", "P1.Summary.Table.rds")
+summarytable_file <- here::here("products","supplement", "P1.Summary.Table.rds")
 saveRDS(P1summary_df, file = summarytable_file)
@@
 ```
-
-```{r ALL Data fit and stat analysis}
-
-```
-######################################
-#Data fitting/statistical analysis
-######################################
-```{r}
-# fit glm.
-# Ct~X= odds of Ct as response variable
-P1.glm.fit<-glm(Ct~Dilution+Detector+Method, data=P1)
-#look at fit
-print(tidy(P1.glm.fit))
-
-step(P1.glm.fit, test="LRT")
-```
-```{r}
-# save fit results table  
-table_file = here("results", "resulttable.rds")
-saveRDS(lmtable, file = table_file)
-
-```

### 2021-11-14 code/analysis_code/03-P2.Data.rmd
Adjust and re-knit analysis script. Re-do modeling still working on it

--- 2021-11-04 code/analysis_code/03-P2.Data.rmd
@@
 P2summary_df <- data.frame(do.call(cbind, lapply(P2, summary)))
+
+print(P2summary_df)
 ```
 ```{r}
-
+P2summary_df<-P2%>%
+  filter(Dilution!="NA", Tool!="SD", Tool!="*.CTRL")%>%
+  mutate(
+    Ct.Cat = ifelse(Ct >=38,"Detected","Positive"),
+    Ct.Cat=factor(Ct.Cat))%>%
+  group_by(Dilution,Method,Ct.Cat)%>%
+  summarise(n=n(),
+            mean=mean(Ct))%>%
+  arrange(Dilution,mean)
 ```
@@
 #save data frame table to file for later use in manuscript
-summarytable_file <- here::here("products","Project_Parts","Part_3", "P2.Summary.Table.rds")
+summarytable_file <- here::here("products","supplement","P2.Summary.Table.rds")
 saveRDS(P2summary_df, file = summarytable_file)
@@
 
+```{r Table Best Ct comparing spike and surface recovery}
+#Summary of Tool A compared to -1 Serial Dilution Recovery
+P2.1Dilution.Summary<-P2%>%
+  filter(Dilution=="-1")%>%
+  mutate(
+    Ct.Cat = ifelse(Ct >=38,"Detected","Positive"),
+    Ct.Cat=factor(Ct.Cat))%>%
+  group_by(Tool,Method,Ct.Cat)%>%
+  summarise(n=n(),
+            mean=mean(Ct))%>%
+  arrange(mean)%>%
+  na.omit(C.Cat)
+```
+```{r}
+#save data frame table to file for later use in manuscript
+summarytable_file <- here::here("products","supplement", "P2.-1Dilution.Table.rds")
+saveRDS(P2.1Dilution.Summary, file = summarytable_file)
+```
+A difference of 2 Ct values when compared by methods for each tool type. CDC was better when directly extracted. NTP performed better during a spiked tool and surface recovery procedures. 
+Ct loss between Spike and SS recovery are about 3 Ct values, equivalent to a log loss of copies.
+
 Make a scatterplot of data
 
-```{r}
+```{r warning=FALSE}
 P2P1<-P2 %>% 
@@
 ```
-```{r}
+```{r warning=FALSE}
 #save figure
@@
 ```
-
-```{r ALL Data fit and stat analysis}
-
-```
-######################################
-#Data fitting/statistical analysis
-######################################
-```{r}
-# fit glm.
-# Ct~X= odds of Ct as response variable
-P2.glm.fit<-glm(Ct~Dilution+Detector, data=P2P1)
-#look at fit
-print(tidy(P2.glm.fit))
-
-step(P2.glm.fit, test="LRT")
-```

### 2021-11-19 code/analysis_code/04-Mods.P2.BestTree.rmd
Update Models

---
title: "Best Tree"
author: "MYC"
date: "11/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Helper packages
library(tidyverse) # for data manipulation & graphics suite
library(here)      # for easy specification of relative locations
library(tidymodels)# for modeling process suite  
library(caret)     # for resampling and model training
library(rpart)
library(rpart.plot)
library(glmnet)
```
Path to data
```{r Data}
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","P2-20211029.rds")
#read and load data. 
```
Import P2 data
```{r}
P2<- readRDS(data_location)
```

Add to the data -- Fix so we add a catagorical aspect to Ct values- simplicity make Detected/Yes (< 37.9) and Undetected/No (>38)

Bonus- try to figure out how to change into 3 levels?
```{r}
P2.Plus.Cat<-
  P2%>%
  mutate(
# if greater than ct 38, this is a negative detection, thus less 38 is a positive in detection
    Ct.Cat = ifelse(Ct >=38,"No","Yes"), 
# There are some NAs, they are undetectable, change only in the Ct.Cat to No as a factor    
    Ct.Cat=ifelse(is.na(Ct.Cat), "No",Ct.Cat),
# These are factors so mutate the chr to fctrs
    Ct.Cat=factor(Ct.Cat)
    )%>%
#Remove/adjust so there are no character based factors
    select(PCR, RH,TEMP, Method, Tool, Input, Detector, Ct, Ct.Cat)%>%
  na.omit()

summary(P2.Plus.Cat)  
```

Set Seed for reproducibility
```{r}
# Fix the random numbers by setting the seed 
set.seed(123)
# This enables the analysis to be reproducible when random numbers are used
```

1. Data splitting into training and testing sets (Random sampling)
  1. Training
  2. Testing

```{r}
# Put 3/4 of the data into the training set; this leaves 1/4 of the data to be used to test 
data_split <- initial_split(P2.Plus.Cat, prop = 3/4)

## Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```

```{r}
#View the train and test data
glimpse(train_data)
glimpse(test_data)
```

2. Creating Models

#  Ct value or Detection as function of method and inital dilution 
Detection ~ Model + Dilution

# Ct value or detection for all precitcors (short hand for all predictors)
Detection~., data=P2.Plus.Cat

2.1. Many engines Examples
```{r}
# fit linear model
lm_lm    <- lm(Ct ~ ., data = P2.Plus.Cat)
# fit general linear model
lm_glm   <- glm(Ct ~ ., data = P2.Plus.Cat, 
                family = gaussian)
#########
#<atempt to use later>
# meta engine (aggregator) method=<method name>; see parsnips
#lm_caret <- train(Ct ~ ., data = P2.Plus.Cat, 
#                 method = "glm")
```

2.4. Resampling - CV
```{r }
# create CV object from training data
cv_data <- rsample::vfold_cv(train_data, v = 5, repeats = 5, strata = 'Ct')

print(cv_data)

############## below are similar functions but need to try

#apply CV in ML algorithm to process the ML model to each resample
vfold_cv(P2.Plus.Cat, v=5)

#bootstrapping- random sample of data with replacment
bootstraps(P2.Plus.Cat, v=5)
```
Preprocessing
Using parsnims to create a recipe for fitting the model
No standardization
adding dummy variables are added
```{r}
fit_recipe <- 
# set a recipe
  recipe(Ct ~ ., data = P2.Plus.Cat) %>%
# code all categorical variables as dummy variables
  step_dummy(all_nominal())

print(fit_recipe)
```
Null Model
Ct is a continuions outcome of numbers-- need to use RMSE as performance metric.
```{r}
RMSE_null_train <- sqrt(sum( (train_data$Ct - mean(train_data$Ct))^2 )/nrow(train_data))

RMSE_null_test <- sqrt(sum( (test_data$Ct - mean(test_data$Ct))^2 )/nrow(test_data))

print(RMSE_null_train)

#predict(lm_lm, train_data=Ct)
```

2.5. Hyper parameter tuning
```{r}

```

2.6. Model evaluation
     - Regression
       - minimize
         - MSE
         - RMSE
         - Deviance
         - MAE
         - RMSLE
       - maximize
         - R^2
     - classification
       - minimize
         - Misclassifcation
         - Mean per class e rror
         - MSE
         - Cross-entropy -- log loss/deviance
         - Gini index
       - maximize
         - Accuracy
         - Precision
         - Sensitivity / recall
         - Specificity
         - AUC

```{r}

```

Tree model
```{r}
tree_model <-  decision_tree() %>% 
  set_args( cost_complexity = tune(), tree_depth = tune(), min_n = tune()) %>% 
  set_engine("rpart") %>% 
  set_mode("regression")      
```
Workflow
```{r}
#set workflow
tree_wf <- workflow() %>%
  add_model(tree_model) %>%
  add_recipe(fit_recipe)
```
Tree Model
```{r}
#tuning grid
tree_grid <- dials::grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)
#tune the model
tree_tune_res <- tree_wf %>% 
  tune::tune_grid(
    resamples = cv_data,
    grid = tree_grid,
    metrics = yardstick::metric_set(rmse) 
  )
```

```{r}
#see a plot of performance for different tuning parameters
tree_tune_res %>% autoplot()
```

```{r}
# get the tuned model that performs best 
best_tree <- tree_tune_res %>%  select_best(metric = "rmse")
# finalize workflow with best model
best_tree_wf <- tree_wf %>% finalize_workflow(best_tree)
# fitting best performing model
best_tree_fit <- best_tree_wf %>% 
                 fit(data = train_data)
#predicting outcomes for final model
tree_pred <- predict(best_tree_fit, train_data)
```

```{r}
rpart.plot(extract_fit_parsnip(best_tree_fit)$fit)
```

```{r}
#predicted versus observed
plot(tree_pred$.pred,train_data$Ct, xlim =c(0,100), ylim=c(0,100))
abline(a=0,b=1, col = 'red') #45 degree line, along which the results should fall

```

```{r}
#residuals
plot(tree_pred$.pred-train_data$Ct)
abline(a=0,b=0, col = 'red') #straight line, along which the results should fall
```

```{r}
tree_perfomance <- tree_tune_res %>% show_best(n = 5)
print(tree_perfomance)
```

### 2021-11-19 code/analysis_code/04-Mods.P2.LASSO.rmd
Update Models

--- 2021-11-19 code/analysis_code/04-Mods.P2.BestTree.rmd
@@
 ---
-title: "Best Tree"
+title: "LASSO"
 author: "MYC"
@@
 
-2.5. Hyper parameter tuning
+# LASSO Linear model
+
+## Set up
 ```{r}
+#model
+lasso_model <- linear_reg() %>%
+  set_mode("regression") %>%           
+  set_engine("glmnet") %>%
+  set_args(penalty = tune(), mixture = 1) #mixture = 1 means we use the LASSO model
 
+#workflow
+lasso_wf <- workflow() %>%
+  add_model(lasso_model) %>% 
+  add_recipe(fit_recipe)
 ```
 
-2.6. Model evaluation
-     - Regression
-       - minimize
-         - MSE
-         - RMSE
-         - Deviance
-         - MAE
-         - RMSLE
-       - maximize
-         - R^2
-     - classification
-       - minimize
-         - Misclassifcation
-         - Mean per class e rror
-         - MSE
-         - Cross-entropy -- log loss/deviance
-         - Gini index
-       - maximize
-         - Accuracy
-         - Precision
-         - Sensitivity / recall
-         - Specificity
-         - AUC
-
-```{r}
-
-```
-
-Tree model
-```{r}
-tree_model <-  decision_tree() %>% 
-  set_args( cost_complexity = tune(), tree_depth = tune(), min_n = tune()) %>% 
-  set_engine("rpart") %>% 
-  set_mode("regression")      
-```
-Workflow
-```{r}
-#set workflow
-tree_wf <- workflow() %>%
-  add_model(tree_model) %>%
-  add_recipe(fit_recipe)
-```
-Tree Model
+## Tuning
 ```{r}
 #tuning grid
-tree_grid <- dials::grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)
-#tune the model
-tree_tune_res <- tree_wf %>% 
-  tune::tune_grid(
-    resamples = cv_data,
-    grid = tree_grid,
-    metrics = yardstick::metric_set(rmse) 
+lasso_reg_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30))
+
+#tune model
+lasso_tune_res <- lasso_wf %>% 
+  tune_grid(resamples = cv_data,
+            grid = lasso_reg_grid,
+            control = control_grid(save_pred = TRUE),
+            metrics = metric_set(rmse)
             )
@@
 
+## Evaluation
+
 ```{r}
 #see a plot of performance for different tuning parameters
-tree_tune_res %>% autoplot()
+lasso_tune_res %>% autoplot()
+```
+
+#
+```{r}
+# get the tuned model that performs best 
+best_lasso <- lasso_tune_res %>%  select_best(metric = "rmse")
+
+# finalize workflow with best model
+best_lasso_wf <- lasso_wf %>% finalize_workflow(best_lasso)
+
+# fitting best performing model
+best_lasso_fit <- best_lasso_wf %>% 
+  fit(data = train_data)
+lasso_pred <- predict(best_lasso_fit, train_data)
 ```
@@
 ```{r}
-# get the tuned model that performs best 
-best_tree <- tree_tune_res %>%  select_best(metric = "rmse")
-# finalize workflow with best model
-best_tree_wf <- tree_wf %>% finalize_workflow(best_tree)
-# fitting best performing model
-best_tree_fit <- best_tree_wf %>% 
-                 fit(data = train_data)
-#predicting outcomes for final model
-tree_pred <- predict(best_tree_fit, train_data)
+x <- best_lasso_fit$fit$fit$fit
+plot(x, "lambda")
+```
+Variables that are part of the best-fit LASSO model, i.e. those that have a non-zero coefficient.
+```{r}
+tidy(extract_fit_parsnip(best_lasso_fit)) %>% filter(estimate != 0)
 ```
 
-```{r}
-rpart.plot(extract_fit_parsnip(best_tree_fit)$fit)
-```
-
+Plotting observed/predicted and residuals.
 ```{r}
 #predicted versus observed
-plot(tree_pred$.pred,train_data$Ct, xlim =c(0,100), ylim=c(0,100))
+plot(lasso_pred$.pred,train_data$Ct, xlim =c(15,50), ylim=c(15,50))
 abline(a=0,b=1, col = 'red') #45 degree line, along which the results should fall
-
 ```
-
 ```{r}
 #residuals
-plot(tree_pred$.pred-train_data$Ct)
+plot(lasso_pred$.pred-train_data$Ct)
 abline(a=0,b=0, col = 'red') #straight line, along which the results should fall
 ```
+Looks decent, the points are along the red lines in the plot
 
+## Model Performance
 ```{r}
-tree_perfomance <- tree_tune_res %>% show_best(n = 5)
-print(tree_perfomance)
+lasso_perfomance <- lasso_tune_res %>% show_best(n = 5)
+print(lasso_perfomance)
 ```

### 2021-11-19 code/analysis_code/04-Mods.P2.RandomForest.rmd
Update Models

--- 2021-11-19 code/analysis_code/04-Mods.P2.LASSO.rmd
@@
 ---
-title: "LASSO"
+title: "Random Forest"
 author: "MYC"
@@
 
-# LASSO Linear model
+```{r}
+rf_model <- rand_forest() %>%
+  set_args(mtry = tune(),     
+    trees = tune(),
+    min_n = tune()
+  ) %>%
+  # select the engine/package that underlies the model
+  set_engine("ranger",
+             num.threads = 18, #for some reason for RF, we need to set this in the engine too
+             importance = "permutation") %>%
+  # choose either the continuous regression or binary classification mode
+  set_mode("regression")   
+```
 
-## Set up
 ```{r}
-#model
-lasso_model <- linear_reg() %>%
-  set_mode("regression") %>%           
-  set_engine("glmnet") %>%
-  set_args(penalty = tune(), mixture = 1) #mixture = 1 means we use the LASSO model
-
 #workflow
-lasso_wf <- workflow() %>%
-  add_model(lasso_model) %>% 
+rf_wf <- workflow() %>%
+  add_model(rf_model) %>% 
   add_recipe(fit_recipe)
@@
 
-## Tuning
 ```{r}
 #tuning grid
-lasso_reg_grid <- tibble(penalty = 10^seq(-3, 0, length.out = 30))
+rf_grid  <- expand.grid(mtry = c(3, 4, 5, 6), min_n = c(40,50,60), trees = c(500,1000)  )
 
-#tune model
-lasso_tune_res <- lasso_wf %>% 
-  tune_grid(resamples = cv_data,
-            grid = lasso_reg_grid,
-            control = control_grid(save_pred = TRUE),
+# tune the model, optimizing RMSE
+rf_tune_res <- rf_wf %>%
+  tune_grid(
+            resamples = cv_data, #CV object
+            grid = rf_grid, # grid of values to try
             metrics = metric_set(rmse) 
@@
 
-## Evaluation
-
 ```{r}
 #see a plot of performance for different tuning parameters
-lasso_tune_res %>% autoplot()
+rf_tune_res %>% autoplot()
 ```
-
-#
 ```{r}
 # get the tuned model that performs best 
-best_lasso <- lasso_tune_res %>%  select_best(metric = "rmse")
+best_rf <- rf_tune_res %>%  select_best(metric = "rmse")
 
 # finalize workflow with best model
-best_lasso_wf <- lasso_wf %>% finalize_workflow(best_lasso)
+best_rf_wf <- rf_wf %>% finalize_workflow(best_rf)
 
 # fitting best performing model
-best_lasso_fit <- best_lasso_wf %>% 
+best_rf_fit <- best_rf_wf %>% 
   fit(data = train_data)
-lasso_pred <- predict(best_lasso_fit, train_data)
+rf_pred <- predict(best_rf_fit, train_data)
 ```
@@
 ```{r}
-x <- best_lasso_fit$fit$fit$fit
-plot(x, "lambda")
+#pull out the fit object
+x <- best_rf_fit$fit$fit$fit
+#plot variable importance
+vip::vip(x, num_features = 20)
 ```
-Variables that are part of the best-fit LASSO model, i.e. those that have a non-zero coefficient.
-```{r}
-tidy(extract_fit_parsnip(best_lasso_fit)) %>% filter(estimate != 0)
-```
+Meaning that the point of Tool B is the most important/predicitive variable for Ct outcome
 
-Plotting observed/predicted and residuals.
 ```{r}
 #predicted versus observed
-plot(lasso_pred$.pred,train_data$Ct, xlim =c(15,50), ylim=c(15,50))
+plot(rf_pred$.pred,train_data$Ct, xlim =c(10,50), ylim=c(10,50))
 abline(a=0,b=1, col = 'red') #45 degree line, along which the results should fall
@@
 #residuals
-plot(lasso_pred$.pred-train_data$Ct)
+plot(rf_pred$.pred-train_data$Ct)
 abline(a=0,b=0, col = 'red') #straight line, along which the results should fall
 ```
-Looks decent, the points are along the red lines in the plot
-
-## Model Performance
 ```{r}
-lasso_perfomance <- lasso_tune_res %>% show_best(n = 5)
-print(lasso_perfomance)
+rf_perfomance <- rf_tune_res %>% show_best(n = 5)
+print(rf_perfomance)
 ```

### 2021-11-24 code/analysis_code/05-Mods.P2.TargetedML.rmd
Pre holiday Push
Push current updates to see what else can be done  for manuscript

---
title: "Random Forest- Targeted"
author: "MYC"
date: "11/24/2021"
output: html_document
---
Interesting results from the P2 Random Forest ML application, Try more targeted variables 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Helper packages
library(tidyverse) # for data manipulation & graphics suite
library(here)      # for easy specification of relative locations
library(tidymodels)# for modeling process suite  
library(caret)     # for resampling and model training
library(rpart)
library(rpart.plot)
library(glmnet)
```
Path to data
```{r Data}
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","P2-20211029.rds")
#read and load data. 
```
Import P2 data
```{r}
P2<- readRDS(data_location)
```

Add to the data -- Fix so we add a categorical aspect to Ct values- simplicity make Detected/Yes (< 37.9) and Undetected/No (>38)

Compare only dilutions that were -1.

```{r}
P2.minus1<-
  P2%>%
  filter(Tool!="B")%>% #remove B because this is actually a -3 dilution. Looking specifically at -1, keep SD range. 
  mutate(
# if greater than ct 38, this is a negative detection, thus less 38 is a positive in detection
    Ct.Cat = ifelse(Ct >=38,"No","Yes"), 
# There are some NAs, they are undetectable, change only in the Ct.Cat to No as a factor    
    Ct.Cat=ifelse(is.na(Ct.Cat), "No",Ct.Cat),
# These are factors so mutate the chr to fctrs
    Ct.Cat=factor(Ct.Cat)
    )%>%
#Remove/adjust so there are no character based factors
    select(PCR, RH,TEMP, Method, Tool, Input, Detector, Ct, Ct.Cat,Dilution)%>%
  na.omit()

summary(P2.minus1)  
```

Set Seed for reproducibility
```{r}
# Fix the random numbers by setting the seed 
set.seed(123)
# This enables the analysis to be reproducible when random numbers are used
```

1. Data splitting into training and testing sets (Random sampling)
  1. Training
  2. Testing

```{r}
# Put 3/4 of the data into the training set; this leaves 1/4 of the data to be used to test 
data_split <- initial_split(P2.minus1, prop = 3/4)

## Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```

```{r}
#View the train and test data
glimpse(train_data)
glimpse(test_data)
```

2. Creating Models

#  Ct value or Detection as function of method and inital dilution 
Detection ~ Method + Dilution

# Ct value or detection for all precitcors (short hand for all predictors)
Detection~., data=P2.minus1

2.1. Many engines Examples
```{r}
# fit linear model
lm_lm    <- lm(Ct ~ ., data = P2.minus1)
# fit general linear model
lm_glm   <- glm(Ct ~ ., data = P2.minus1, 
                family = gaussian)
#########
#<atempt to use later>
# meta engine (aggregator) method=<method name>; see parsnips
#lm_caret <- train(Ct ~ ., data = P2.Plus.Cat, 
#                 method = "glm")
```

2.4. Resampling - CV
```{r }
# create CV object from training data
cv_data <- rsample::vfold_cv(train_data, v = 5, repeats = 5, strata = 'Ct')

print(cv_data)

############## below are similar functions but need to try

#apply CV in ML algorithm to process the ML model to each resample
vfold_cv(P2.minus1, v=5)

#bootstrapping- random sample of data with replacment
bootstraps(P2.minus1, v=5)
```
Preprocessing
Using parsnips to create a recipe for fitting the model
No standardization
adding dummy variables are added
```{r}
fit_recipe <- 
# set a recipe
  recipe(Ct ~., data = P2.minus1) %>%
# code all categorical variables as dummy variables
  step_dummy(all_nominal())

print(fit_recipe)
```
Null Model
Ct is a continuions outcome of numbers-- need to use RMSE as performance metric.
```{r}
RMSE_null_train <- sqrt(sum( (train_data$Ct - mean(train_data$Ct))^2 )/nrow(train_data))

RMSE_null_test <- sqrt(sum( (test_data$Ct - mean(test_data$Ct))^2 )/nrow(test_data))

print(RMSE_null_train)

#predict(lm_lm, train_data=Ct)
```

```{r}
rf_model <- rand_forest() %>%
  set_args(mtry = tune(),     
    trees = tune(),
    min_n = tune()
  ) %>%
  # select the engine/package that underlies the model
  set_engine("ranger",
             num.threads = 18, #for some reason for RF, we need to set this in the engine too
             importance = "permutation") %>%
  # choose either the continuous regression or binary classification mode
  set_mode("regression")   
```

```{r}
#workflow
rf_wf <- workflow() %>%
  add_model(rf_model) %>% 
  add_recipe(fit_recipe)
```

```{r}
#tuning grid
rf_grid  <- expand.grid(mtry = c(3, 4, 5, 6), min_n = c(40,50,60), trees = c(500,1000)  )

# tune the model, optimizing RMSE
rf_tune_res <- rf_wf %>%
  tune_grid(
            resamples = cv_data, #CV object
            grid = rf_grid, # grid of values to try
            metrics = metric_set(rmse) 
  )
```

```{r}
#see a plot of performance for different tuning parameters
rf_tune_res %>% autoplot()
```
```{r}
# get the tuned model that performs best 
best_rf <- rf_tune_res %>%  select_best(metric = "rmse")

# finalize workflow with best model
best_rf_wf <- rf_wf %>% finalize_workflow(best_rf)

# fitting best performing model
best_rf_fit <- best_rf_wf %>% 
  fit(data = train_data)
rf_pred <- predict(best_rf_fit, train_data)
```

```{r}
#pull out the fit object
x <- best_rf_fit$fit$fit$fit
#plot variable importance
P2RFBestFitminus1<-vip::vip(x, num_features = 20)
print(P2RFBestFitminus1)
```
Interesting so Dilution is obviously more important followed by Input. It's more surprizing that the Method is one of the least important in this comparison. Infact the temperature and relative humidity are more impractically thatn the method.
```{r warning=FALSE}
#save figure
figure_file = here("results","P2-RF-BestFit-minus1.png")
ggsave(filename = figure_file, plot=P2RFBestFitminus1) 
```

```{r}
#predicted versus observed
plot(rf_pred$.pred,train_data$Ct, xlim =c(10,50), ylim=c(10,50))
abline(a=0,b=1, col = 'red') #45 degree line, along which the results should fall
```
```{r}
#residuals
plot(rf_pred$.pred-train_data$Ct)
abline(a=0,b=0, col = 'red') #straight line, along which the results should fall
```
```{r}
rf_perfomance <- rf_tune_res %>% show_best(n = 5)
print(rf_perfomance)
```

# Try to see if Best Tree can show us something

Tree model
```{r}
tree_model <-  decision_tree() %>% 
  set_args( cost_complexity = tune(), tree_depth = tune(), min_n = tune()) %>% 
  set_engine("rpart") %>% 
  set_mode("regression")      
```
Workflow
```{r}
#set workflow
tree_wf <- workflow() %>%
  add_model(tree_model) %>%
  add_recipe(fit_recipe)
```
Tree Model
```{r}
#tuning grid
tree_grid <- dials::grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)
#tune the model
tree_tune_res <- tree_wf %>% 
  tune::tune_grid(
    resamples = cv_data,
    grid = tree_grid,
    metrics = yardstick::metric_set(rmse) 
  )
```

```{r}
#see a plot of performance for different tuning parameters
tree_tune_res %>% autoplot()
```
```{r}
# get the tuned model that performs best 
best_tree <- tree_tune_res %>%  select_best(metric = "rmse")
# finalize workflow with best model
best_tree_wf <- tree_wf %>% finalize_workflow(best_tree)
# fitting best performing model
best_tree_fit <- best_tree_wf %>% 
                 fit(data = train_data)
#predicting outcomes for final model
tree_pred <- predict(best_tree_fit, train_data)
```

```{r}
rpart.plot(extract_fit_parsnip(best_tree_fit)$fit)
```

### 2021-11-24 products/manuscript/Manuscript.Rmd
Pre holiday Push
Push current updates to see what else can be done  for manuscript

@@
 
-Applying MADA learned techniques to a RT-PCR dataset to determine the effectiveness of sample enhancement strategies for environmental surface sampling of low bio-burden viruses.
+Applying MADA learned techniques to a RT-PCR data set to determine the effectiveness of sample enhancement strategies for environmental surface sampling of low bio-burden viruses.
 
@@
 ## Background
+
 The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants.
 Environmental reservoirs are a reported, indirect transmission route for viral pathogens [@Marcenac2021-sz].
-Although many viruses have a low infectious dose, they can be difficult to detect due to the low bioburden in the environment.
+Although many viruses have a low infectious dose, they can be difficult to detect due to the low bio-burden in the environment.
 The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control [@Paton2021-jc].
 Culture-based methods can be labor and time intensive.
-The use and limitations of commercial “target capture” (TC) technologies are varied compared to molecular approaches utilizing automated extraction.
-The objective of this study is improve outbreak response through rapid pathogen detection by evaluating a TC technology’s ability to capture SC2 from complex matrices.
+The use and limitations of commercial "target capture" (TC) technologies are varied compared to molecular approaches utilizing automated extraction.
+The objective of this study is improve outbreak response through rapid pathogen detection by evaluating a TC technology's ability to capture SC2 from complex matrices.
 
@@
 ## Data aquisition
+
 Reports per each rRT-PCR run were generated and quality checked on their respective machines before transfer to a Teams repository.
@@
 
-Data was imported and cleaned.
-Data was then split between the two different experimental types (P1 and P2).
+Data was imported and cleaned. Data was then split between the two different experimental types (P1 and P2).
 
@@
 
-Atempt to do as [@Brandolini2021-ja]
-
 #### All Data
@@
 
-Summary table 
-scatter plot of SDs based on dilution factors
+Summary table scatter plot of SDs based on dilution factors
+
 ```{r}
@@
 ```
+
 #### Phase 1
@@
 ```
+
 #### Phase 2 Comparing A and B Tool Recovery
+
 ```{r}
@@
 ```
+
 Tool A has an average Ct recovery higher than B.
 This makes sense as the surfaces A recovered from is 2 logs higher (-1) than that of B (-3).
-A log drop is ~3Ct values we can see from the graphic that it trends as predicted.
+A log drop is \~3Ct values we can see from the graphic that it trends as predicted.
 Suggests that
@@
 #### Phase 2 Comparing Method Serial Dilutions (CDC vs NTP) with standard input of 300uL
+
 ```{r}
@@
 ```
+
 Serial dilution data between CDC and NTP.
@@
 #### Phase 2 Comparing Methods with standard input of 300uL
+
 ```{r}
@@
 #### Phase 2 Method extraction comparisons with all inputs and tools
+
 ```{r}
@@
 #### Phase 2 Nanoparticles only - Compare detectors with dilution and input in Serial Dilutions
+
 ```{r}
@@
 #### Phase 2 Nanoparticles only - All tools at 300 and 1000ul input levels
+
 ```{r}
@@
 ## Summary and Interpretation
+
 Took
@@
 ## Strengths and Limitations
+
 Need more replicates, but this is a time consuming processes just to get the data.

### 2021-11-28 code/analysis_code/04-Mods.P2.RandomForest.rmd
Post Holiday Push
Push of analysis completion. Loosing internet 11/28 and will update coherent manuscript edits by Monday Due date.

@@
 #Remove/adjust so there are no character based factors
-    select(PCR, RH,TEMP, Method, Tool, Input, Detector, Ct, Ct.Cat)%>%
+    select(RH,TEMP, Method, Tool, Input, Detector, Ct, Ct.Cat)%>%
   na.omit()
@@
 
-#bootstrapping- random sample of data with replacment
-bootstraps(P2.Plus.Cat, v=5)
 ```
@@
 Null Model
-Ct is a continuions outcome of numbers-- need to use RMSE as performance metric.
+Ct is a continuations outcome of numbers-- need to use RMSE as performance metric.
 ```{r}
@@
 ```
+Model tuning with a grid.
 
@@
 ```
-
+Plot the performance of the grid tune results.
 ```{r}
@@
 ```
+
 ```{r}
@@
 ```
-
+Use vip to learn about variable importance in the modeling
 ```{r}
@@
 #plot variable importance
-vip::vip(x, num_features = 20)
+P2RFBestFit<-vip::vip(x, geom="point",aesthetics = list(color = "blue", shape = 17, size = 4))
+
+print(P2RFBestFit)
 ```
-Meaning that the point of Tool B is the most important/predicitive variable for Ct outcome
+Meaning that the point of Tool B is the most important/predictive variable for Ct outcome, makes sense as the top 3 are tool related and all have different "concentrations".
+Also not pressurizing that Dilution and Input are important.
+Method has virtually 0 importance in the Ct value during the modeling. 
+```{r warning=FALSE}
+#save figure
+figure_file = here("results","P2-RF-BestFit.png")
+ggsave(filename = figure_file, plot=P2RFBestFit) 
+```
 
@@
 #predicted versus observed
-plot(rf_pred$.pred,train_data$Ct, xlim =c(10,50), ylim=c(10,50))
+plot(rf_pred$.pred,train_data$Ct )
 abline(a=0,b=1, col = 'red') #45 degree line, along which the results should fall

### 2021-11-28 code/analysis_code/05-Mods.P2.TargetedML.rmd
Post Holiday Push
Push of analysis completion. Loosing internet 11/28 and will update coherent manuscript edits by Monday Due date.

@@
 #Remove/adjust so there are no character based factors
-    select(PCR, RH,TEMP, Method, Tool, Input, Detector, Ct, Ct.Cat,Dilution)%>%
+    select(RH,TEMP, Method, Tool, Input, Detector, Ct, Ct.Cat,Dilution)%>%
   na.omit()
@@
 # Put 3/4 of the data into the training set; this leaves 1/4 of the data to be used to test 
-data_split <- initial_split(P2.minus1, prop = 3/4)
+data_split2 <- initial_split(P2.minus1, prop = 3/4)
 
 ## Create data frames for the two sets:
-train_data <- training(data_split)
-test_data  <- testing(data_split)
+train_data2 <- training(data_split2)
+test_data2  <- testing(data_split2)
 ```
@@
 #View the train and test data
-glimpse(train_data)
-glimpse(test_data)
+glimpse(train_data2)
+glimpse(test_data2)
+
+table(train_data2$Method)%>%
+  prop.table()
+table(test_data2$Method)%>%
+  prop.table()
 ```
@@
 # meta engine (aggregator) method=<method name>; see parsnips
-#lm_caret <- train(Ct ~ ., data = P2.Plus.Cat, 
-#                 method = "glm")
+lm_caret <- train(Ct ~ Method + Dilution + Input, data = P2.minus1, 
+                 method = "glm")
 ```
@@
 ```{r }
-# create CV object from training data
-cv_data <- rsample::vfold_cv(train_data, v = 5, repeats = 5, strata = 'Ct')
-
-print(cv_data)
-
-############## below are similar functions but need to try
-
-#apply CV in ML algorithm to process the ML model to each resample
-vfold_cv(P2.minus1, v=5)
-
-#bootstrapping- random sample of data with replacment
-bootstraps(P2.minus1, v=5)
+# create bootstrap resample object from training data
+cv_data_boots <- bootstraps(train_data2, v=5)
+print(cv_data_boots)
 ```
@@
 ```{r}
-fit_recipe <- 
-# set a recipe
-  recipe(Ct ~., data = P2.minus1) %>%
+fit_recipe2 <- 
+# set a recipe specific to the Dilution and Input
+  recipe(Ct ~Dilution+Input+Method, data = P2.minus1) %>%
 # code all categorical variables as dummy variables
@@
 
-print(fit_recipe)
+print(fit_recipe2)
 ```
@@
 ```{r}
-RMSE_null_train <- sqrt(sum( (train_data$Ct - mean(train_data$Ct))^2 )/nrow(train_data))
+RMSE_null_train2 <- sqrt(sum( (train_data2$Ct - mean(train_data2$Ct))^2 )/nrow(train_data2))
 
-RMSE_null_test <- sqrt(sum( (test_data$Ct - mean(test_data$Ct))^2 )/nrow(test_data))
+RMSE_null_test2 <- sqrt(sum( (test_data2$Ct - mean(test_data2$Ct))^2 )/nrow(test_data2))
 
-print(RMSE_null_train)
+print(RMSE_null_train2)
 
-#predict(lm_lm, train_data=Ct)
+predict(lm_glm, train_data=Ct)
 ```
@@
 ```{r}
-rf_model <- rand_forest() %>%
+rf_model2 <- rand_forest() %>%
   set_args(mtry = tune(),     
@@
 #workflow
-rf_wf <- workflow() %>%
-  add_model(rf_model) %>% 
-  add_recipe(fit_recipe)
+rf_wf2 <- workflow() %>%
+  add_model(rf_model2) %>% 
+  add_recipe(fit_recipe2)
 ```
-
+Try bootstrapping in CV workflow
 ```{r}
 #tuning grid
-rf_grid  <- expand.grid(mtry = c(3, 4, 5, 6), min_n = c(40,50,60), trees = c(500,1000)  )
+rf_grid2  <- expand.grid(mtry = c(3, 4, 5, 6), min_n = c(40,50,60), trees = c(500,1000)  )
 
 # tune the model, optimizing RMSE
-rf_tune_res <- rf_wf %>%
+rf_tune_res2 <- rf_wf2 %>%
   tune_grid(
-            resamples = cv_data, #CV object
-            grid = rf_grid, # grid of values to try
+            resamples = cv_data_boots, #CV object
+            grid = rf_grid2, # grid of values to try
             metrics = metric_set(rmse) 
@@
 #see a plot of performance for different tuning parameters
-rf_tune_res %>% autoplot()
+rf_tune_res2 %>% autoplot()
 ```
@@
 # get the tuned model that performs best 
-best_rf <- rf_tune_res %>%  select_best(metric = "rmse")
+best_rf2 <- rf_tune_res2 %>%  select_best(metric = "rmse")
 
 # finalize workflow with best model
-best_rf_wf <- rf_wf %>% finalize_workflow(best_rf)
+best_rf_wf2 <- rf_wf2 %>% finalize_workflow(best_rf2)
 
 # fitting best performing model
-best_rf_fit <- best_rf_wf %>% 
-  fit(data = train_data)
-rf_pred <- predict(best_rf_fit, train_data)
+best_rf_fit2 <- best_rf_wf2 %>% 
+  fit(data = train_data2)
+rf_pred2 <- predict(best_rf_fit2, train_data2)
 ```
@@
 #pull out the fit object
-x <- best_rf_fit$fit$fit$fit
+x2 <- best_rf_fit2$fit$fit$fit
 #plot variable importance
-P2RFBestFitminus1<-vip::vip(x, num_features = 20)
+P2RFBestFitminus1<-vip::vip(x2, num_features = 10, geom="point", aesthetics = list(color = "red", shape = 17, size = 4))
 print(P2RFBestFitminus1)
@@
 #predicted versus observed
-plot(rf_pred$.pred,train_data$Ct, xlim =c(10,50), ylim=c(10,50))
+plot(rf_pred2$.pred,train_data2$Ct)
 abline(a=0,b=1, col = 'red') #45 degree line, along which the results should fall
@@
 #residuals
-plot(rf_pred$.pred-train_data$Ct)
+plot(rf_pred2$.pred-train_data2$Ct)
 abline(a=0,b=0, col = 'red') #straight line, along which the results should fall
@@
 ```{r}
-rf_perfomance <- rf_tune_res %>% show_best(n = 5)
-print(rf_perfomance)
+rf_perfomance2 <- rf_tune_res2 %>% show_best(n = 5)
+print(rf_perfomance2)
 ```
@@
 ```{r}
-tree_model <-  decision_tree() %>% 
+tree_model2 <-  decision_tree() %>% 
   set_args( cost_complexity = tune(), tree_depth = tune(), min_n = tune()) %>% 
@@
 #set workflow
-tree_wf <- workflow() %>%
-  add_model(tree_model) %>%
-  add_recipe(fit_recipe)
+tree_wf2 <- workflow() %>%
+  add_model(tree_model2) %>%
+  add_recipe(fit_recipe2)
 ```
@@
 #tuning grid
-tree_grid <- dials::grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)
+tree_grid2 <- dials::grid_regular(cost_complexity(), tree_depth(), min_n(), levels = 5)
 #tune the model
-tree_tune_res <- tree_wf %>% 
+tree_tune_res2 <- tree_wf2 %>% 
   tune::tune_grid(
-    resamples = cv_data,
-    grid = tree_grid,
+    resamples = cv_data_boots,
+    grid = tree_grid2,
     metrics = yardstick::metric_set(rmse) 
@@
 #see a plot of performance for different tuning parameters
-tree_tune_res %>% autoplot()
+tree_tune_res2 %>% autoplot()
 ```
@@
 # get the tuned model that performs best 
-best_tree <- tree_tune_res %>%  select_best(metric = "rmse")
+best_tree2 <- tree_tune_res2 %>%  select_best(metric = "rmse")
 # finalize workflow with best model
-best_tree_wf <- tree_wf %>% finalize_workflow(best_tree)
+best_tree_wf2 <- tree_wf2 %>% finalize_workflow(best_tree2)
 # fitting best performing model
-best_tree_fit <- best_tree_wf %>% 
-                 fit(data = train_data)
+best_tree_fit2 <- best_tree_wf2 %>% 
+                 fit(data = train_data2)
 #predicting outcomes for final model
-tree_pred <- predict(best_tree_fit, train_data)
+tree_pred2 <- predict(best_tree_fit2, train_data2)
 ```
@@
 ```{r}
-rpart.plot(extract_fit_parsnip(best_tree_fit)$fit)
+minusTree<-
+  rpart.plot(extract_fit_parsnip(best_tree_fit2)$fit,type = 5, tweak = 1.2)
+
+print(minusTree)
 ```
+Each node shows: predicted values, and percentage of observations in node.
+So if the dilution is still the lead in the effects of the Ct. - Makes sense as that's the spiking/manipulating level 
+If the dilution is greater than -1.5, effects of the input factor strongly in the Ct recovery, with a little input from the method if the input is less than 650uL

### 2021-11-28 products/manuscript/Manuscript.Rmd
Post Holiday Push
Push of analysis completion. Loosing internet 11/28 and will update coherent manuscript edits by Monday Due date.

@@
 bibliography: ../references.bib
-csl: ../apa.csl
+csl: ../PLoS.csl
 ---
@@
 
-Applying MADA learned techniques to a RT-PCR data set to determine the effectiveness of sample enhancement strategies for environmental surface sampling of low bio-burden viruses.
+The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants.
+Environmental reservoirs are a reported, indirect transmission route for viral pathogens.
+Although many viruses have a low infectious dose, they can be difficult to detect due to the low-bioburden in the environment.
+The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control.
+Culture-based methods can be labor and time intensive.
+The use and limitations of commercial “target capture” (TC) technologies are varied compared to molecular approaches utilizing automated extraction.
+The objective of this study is improved outbreak response through rapid pathogen detection by evaluating a TC technology’s ability to capture SC2 from complex matrices.
 
@@
 
+The global emergence of novel human coronavirus, SARS-CoV-2 (SC2) in 2019 became a global health concern.
+SC2 causes severe respiratory tract infections in humans with still unknown future complications.
+With the virus' high infection rate and rapid spreading capabilities via human-to-human transmission
+
 The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants.
-Environmental reservoirs are a reported, indirect transmission route for viral pathogens [@Marcenac2021-sz].
+Environmental reservoirs are a reported, indirect transmission route for viral pathogens [@Tanner2021;@Marcenac2021-sz].
 Although many viruses have a low infectious dose, they can be difficult to detect due to the low bio-burden in the environment.
@@
 
+The virus is capable of human-to-human transmission via inhalation of infected aerosols.
+Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020].
 ### Questions/Hypotheses to be addressed
@@
 
-Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR [@Wittwer2012-vt][@Shah2021-mb].
+Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR [@Wittwer2012-vt;@Shah2021-mb;@Barclay2020].
 
@@
 
-```{r}
+```{r P2-SD, fig.cap="Serial Dilution Comparisons within Phase 2 data", echo=FALSE}
 knitr::include_graphics("../../results/P2-SD-Plot.png")
@@
 
-```{r}
+```{r P2-Tool-AB, echo=FALSE, fig.cap="Phase 2 Comparison of A and B tool recovery"}
 knitr::include_graphics("../../results/ASK-P2P1-Tools.png")
@@
 #### Phase 2 Comparing Method Serial Dilutions (CDC vs NTP) with standard input of 300uL
-
+MOVE TO SUPPLEMENT
 ```{r}
@@
 
-```{r}
+```{r P2Tools-Method, fig.cap="P2 Direct Comparisons of Recovery by Methods used", echo=FALSE}
 knitr::include_graphics("../../results/ASK-P2P3-Methods0300.png")
@@
 
+### Phase 2
+
+Data was limited to Phase 2 data from this point on wards to limit variability in the surface recovery
+
 #### Phase 2 Method extraction comparisons with all inputs and tools
-
-```{r}
-knitr::include_graphics("../../results/ASK-P2P4-MethodsALL.png")
-```
 
@@
 
-```{r}
-knitr::include_graphics("../../results/ASK-P2P5-NTP-SD.png")
-```
-
 #### Phase 2 Nanoparticles only - All tools at 300 and 1000ul input levels
 
-```{r}
+```{r echo=FALSE}
 knitr::include_graphics("../../results/ASK-P2P6-NTP.png")
@@
 ### Data Modeling
+```{r echo=FALSE}
+knitr::include_graphics("../../results/P2-RF-BestFit.png")
+```
+
+```{r ImportanceFactor, fig.cap="Bootstrapped Importance Factor excluding Tools ", echo=FALSE}
+knitr::include_graphics("../../results/P2-RF-BestFit-minus1.png")
+```
+
+```{r minusDecisionTree, fig.cap="Bootstrapped Decision Tree of Limited Factors.", echo=FALSE}
+knitr::include_graphics("../../results/P2-DecisionTree-minus1.png")
+```
 
@@
 
-Figure \@ref(fig:resultfigure) shows a scatterplot figure produced by one of the R scripts.
+Figure \@ref(fig:minusDecisionTree) shows a bootstrapped decision tree figure.
 
@@
 ## Strengths and Limitations
-
+RT-PCR does not distinguish between infectious and non-infectious virus. Additional work will need to be completed to better compare the recovery and estimate the true risk of SC2 in the environment.
 Need more replicates, but this is a time consuming processes just to get the data.
@@
 Additional materials need to be considered [@Jansson2020-cg].
-
+Previous studies have demonstrated capture techniques can overcome limitations of conventional studies, and if applied to environmental samples can yield better data to enlighten the field [@Gasc2018].
 ## Conclusions
+The use of nanoparticles does result in the recovery of Ct values in the assay.
+with the results of the analysis of data there is no significant difference between the automated extraction method and the nanoparticles.
+Using some ML modeling we can see that there are some recorded variables that affect the Ct values- Dilution of the sample and Input.
 

### 2021-11-28 code/analysis_code/04-Mods.P2.BestTree.rmd
Confirmed analysis output knits

@@
 ```{r}
-rpart.plot(extract_fit_parsnip(best_tree_fit)$fit)
+rpart.plot(extract_fit_parsnip(best_tree_fit)$fit, tweak = 1.2)
 ```
@@
 #predicted versus observed
-plot(tree_pred$.pred,train_data$Ct, xlim =c(0,100), ylim=c(0,100))
+plot(tree_pred$.pred,train_data$Ct)
 abline(a=0,b=1, col = 'red') #45 degree line, along which the results should fall

### 2021-11-28 products/manuscript/Manuscript.Rmd
Confirmed analysis output knits

@@
 
-processing script.R was modified to do initial cleaning and preparation of the data set for further exploration.
+`processing script.R` was modified to do initial cleaning and preparation of the data set for further exploration.
 
@@
 
-This can be found in processing_code/ProcessingScript.RMD
+This can be found in `processing_code` folder in repository that generate rds files located in the `data/processed_data` folder.
 
@@
 
-Direction was influenced by [@Weir2016-sj] on "Effect fo Surface Sampling and Recovery of Viruses and No-Spore-Forming Bacteria on a Quantitative Microbial Risk Assessment Model for Fomites".
+Processed data
 
@@
 Suggests that
+```{r P2-Tool-AB-Sum, eval=FALSE, fig.cap="Phase", include=FALSE}
+knitr::kable(table, caption= "Example Table Caption", digits=2, col.names = )
+```
 

### 2021-11-29 products/manuscript/Manuscript.Rmd
Peer Review Push

--- 2021-11-28 products/manuscript/Manuscript.Rmd
@@
 
-## Background
-
-The global emergence of novel human coronavirus, SARS-CoV-2 (SC2) in 2019 became a global health concern.
-SC2 causes severe respiratory tract infections in humans with still unknown future complications.
-With the virus' high infection rate and rapid spreading capabilities via human-to-human transmission
-
 The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants.
@@
 Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020].
-### Questions/Hypotheses to be addressed
 
-There are several major aims of this:
+# Material and methods
+
+## Data collection
+
+### Setup and conditions
+
+Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR.
+[@Wittwer2012-vt;@Shah2021-mb;@Barclay2020].
+Real time RT-PCR was conducted per CDC 2019-nCoV Real-Time RT-PCR Diagnostic Panel protocol to detect the presence of N1 and N2 sequences designed for the specific detection of 2019-nCoV.Reports were QCed and exported from machines electronically.
+
+### Data aquisition and compillation
+
+Summary data was selected from the reports and added to additional environmental data gathered on based on the extraction date of the sample.
+
+A composted data set was compiled and saved as a csv to be used as the base data for the project along with environmental condition data from a specific recovery experiment.
+
+## Data cleaning
+
+Data was cleaned and reformatted to relevant factors and numeric data types.
+During cleaning the data was split and saved in three forms- All (compilation of all recovery studies- P1 and P2);P1 (only data from spiked studies); and P2 (only data from surface sampling studies).
+P1 and P2 were analyzed together due to constant serial dilution controls.
+P2 data were expanded on due to the interests of the study focusing on surface recovery.
+
+## Analysis
+
+Analysis was tailored to three major aims of this study:
 
@@
 3.  Are there any environmental variables recorded that have a positive impact on detection?
-4.  Can this data be modeled to back calculate a sample's original concentration based on the Ct values recovered?
 
-Testing several sample enhancement strategies (Nanoparticles and Bait & Capture) for molecular detection of heat-inactivated SARS-CoV-2 from stainless-steel surfaces.
-The primary aim of this work is to determine if any of the enhancement strategies increase the concentration of RNA copies in a sample that will express itself as a lower Ct value from the Rt-PCR output.
-Lower Ct values equate to more copies being present in the sample meaning less time needed to confirm presence.
+To evaluate Aim 1, All data was utilized to compare Methods and their respective Ct values per serial dilution factor.
+A serial dilution series was utilized across all assays to insure consistencies regardless of study.
 
-In the case the sample enhancements work, we would then compare how well they worked against an automated extraction method.
-The degree of "wellness" will be determined by how low the Ct value results are produced after a sample has going through a specific method.
+For Aim 2, data was limited to P2 data from this point onwards to focus on surface recovery of genomic material and limit variability from P1 spiked studies.
+An exploration in how Tool and Methods differed.
+Additional exploration occured in looking specifically at the NTP Method and its results as there were additional variables in input.
 
-In addition to the comparison of enhancement strategies the effects of the tool used for sampling (macrofoam or knit-fiber), time held, and other environmental variables will be explored any significant effects, with interest in positive impacts to detection [@Jansson2020-cg].
+Final in Aim 3, we implemented several machine learning techniques to determine if it is possible to model Ct recovery with the data gathered and what variables impact the recovered Ct value.
+We utilized the random forest model to see what all factors and their impact and repeated with more specific asks in the model.
+Then we use the best fit tree model to make a decision tree on predictability of recovered Ct Values.
 
-## Methods and Results
+# Results
 
-### Experiment setup and conditions
+## Aim 1: Does NTP work?
 
-Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR [@Wittwer2012-vt;@Shah2021-mb;@Barclay2020].
+We compare Ct values of the serial dilution controls from all studies between CDC and NTP methods.
 
-## Data aquisition
-
-Reports per each rRT-PCR run were generated and quality checked on their respective machines before transfer to a Teams repository.
-Summary data was selected from the reports and added to additional environmental data gathered on based on the extraction date of the sample.
-
-Completed run data is then collated into one file and read into R./A composited data set was compiled and saved as a csv and used as the base data for the project
-
-### Data import and cleaning
-
-processing script.R was modified to do initial cleaning and preparation of the data set for further exploration.
-
-Data was imported and cleaned. Data was then split between the two different experimental types (P1 and P2).
-
-This can be found in processing_code/ProcessingScript.RMD
-
-### Data exploration
-
-Direction was influenced by [@Weir2016-sj] on "Effect fo Surface Sampling and Recovery of Viruses and No-Spore-Forming Bacteria on a Quantitative Microbial Risk Assessment Model for Fomites".
-
-#### All Data
-
-Overview - split between part 1 and part 2
-
-##### Serial Dilutions confirmation
-
-Summary table scatter plot of SDs based on dilution factors
-
-```{r}
+```{r ALL-SD, fig.cap="Serial Dilution Comparisons within Phase 2 data", echo=FALSE}
 knitr::include_graphics("../../results/ALL-SD-Plot.png")
@@
 
-#### Phase 1
+Serial dilution Ct values increase in a linear manner and within expected intervals indicating log10 loss as the dilution value decreases.
+Overall the Methods appear to not be too different from one another at higher dilution levels, but we also see that as the Dilution level decreases the the NTP method registers a lower Ct value than the CDC method and begins to differentiate.
 
-```{r}
-knitr::include_graphics("../../results/P1-SD-Plot.png")
-```
-
-### Phase 2
+## Aim 2: How much more different - Phase 2 Focused Exploration
 
@@
 
-#### Phase 2 Comparing A and B Tool Recovery
+Repeat comparisons of the first Aim but with the data from P2 to see if the differences are evident.
+Overall the Methods appear to not be explicitly different from one another.
+The divergence of CDC and NTP methods is still seen slightly with CDC recovering less at higher dilution levels and switching places with NTP as the dilution increases.
 
-```{r P2-Tool-AB, echo=FALSE, fig.cap="Phase 2 Comparison of A and B tool recovery"}
+### Phase 2 Comparing A and B Tool Recovery
+
+```{r P2-Tool-AB, echo=FALSE, fig.cap="Phase 2 Comparison of A and B tool recovery at different input volumes by Probe"}
 knitr::include_graphics("../../results/ASK-P2P1-Tools.png")
@@
 
+The larger the input volume the recovered Ct values tends to be smaller. 
 Tool A has an average Ct recovery higher than B.
 This makes sense as the surfaces A recovered from is 2 logs higher (-1) than that of B (-3).
-A log drop is \~3Ct values we can see from the graphic that it trends as predicted.
-Suggests that
+A log drop is ~3Ct values we can see from the graphic that it trends as predicted.
+Between methods at an input level of 300ul, there's a lot more variability in the NTP method than the CDC method,.
+Probes see little to no difference in Ct Values recovered.
 
-#### Phase 2 Comparing Method Serial Dilutions (CDC vs NTP) with standard input of 300uL
-MOVE TO SUPPLEMENT
-```{r}
-knitr::include_graphics("../../results/ASK-P2P2-Methods.png")
+### Phase 2 Comparing Methods with standard input of 300uL
+
+```{r P2Tools-Method, fig.cap="P2 Direct Comparisons of Recovery by Methods used", echo=FALSE}
+knitr::include_graphics("../../results/ASK-P2P3-Methods0300.png")
 ```
@@
 
-#### Phase 2 Comparing Methods with standard input of 300uL
-
-```{r P2Tools-Method, fig.cap="P2 Direct Comparisons of Recovery by Methods used", echo=FALSE}
-knitr::include_graphics("../../results/ASK-P2P3-Methods0300.png")
-```
-
-### Phase 2
-
-Data was limited to Phase 2 data from this point on wards to limit variability in the surface recovery
-
-#### Phase 2 Method extraction comparisons with all inputs and tools
-
-#### Phase 2 Nanoparticles only - Compare detectors with dilution and input in Serial Dilutions
-
-#### Phase 2 Nanoparticles only - All tools at 300 and 1000ul input levels
+### Phase 2 Nanoparticles only - All tools at 300 and 1000ul input levels
 
@@
 
-### Data Analysis
+A closer look at specially NTP only results with the CDC input vol and the max vol the manufacturer recommends.
+The increased input volume appears to result in smaller recovery values than the smaller input volume.
+Still no difference between Probe types.
 
-### Data Modeling
-```{r echo=FALSE}
+## Data Modeling
+
+To determine if other variables affect the recovery Ct values.
+
+```{r ImportanceFactor, fig.cap="Random forest modeled Importance Factor- all variables",echo=FALSE}
 knitr::include_graphics("../../results/P2-RF-BestFit.png")
@@
 
-```{r ImportanceFactor, fig.cap="Bootstrapped Importance Factor excluding Tools ", echo=FALSE}
+Random forest model comparing Ct values and all other factors in terms of importance of affects. Top three are all tools. Method and Detector are the least important factors affecting Ct values.
+
+```{r ImportanceFactor2, fig.cap="Bootstrapped Importance Factor excluding Tools", echo=FALSE}
 knitr::include_graphics("../../results/P2-RF-BestFit-minus1.png")
 ```
+
+Limited comparison specifically for Dilution and Input.
+Additionally put in Method to see if there are any affects.
+Dilution plays a larger importance than the input of genomic material in this assay for recoverable Ct values.
+No real importance between method, in fact, negative importance with NTP.
 
@@
 
-## Exploratory analysis
-
-Table \@ref(tab:summarytable) shows a table summarizing the data.
-
-Figure \@ref(fig:minusDecisionTree) shows a bootstrapped decision tree figure.
-
-```{r resultfigure, eval=FALSE, fig.cap='Analysis figure.', include=FALSE}
-knitr::include_graphics("../../results/resultfigure.png")
-```
-
-## Full analysis
-
-Example table \@ref(tab:resulttable) shows a table summarizing a linear model fit.
+Ran the Random Forest parameters as a best tree to create a decision tree.
+We can see that Dilution can be exclusively what determines recovery values when dilutions are less than -1.5.
+If dilutions are above -1.5, the input of sample and method may influence the recovered Ct values.
 
@@
 
-## Summary and Interpretation
+With the linear regression comparison in \@ref(fig:ALL-SD) we see a general trend of increasing Ct value as dilution decreases.
+Interesting to note that within \@ref(fig:ALL-SD) we see a larger divide between methods at the lower dilution level in figure \@ref(fig:P2-SD) as well.
+This suggests that NTP may perform better or be  able to capture the little genomic fragments in a low burdened sample and thus have some targeted/ concentrating abilities.
 
-Took
+In a direct comparison between methods, NTP is more variable than CDC.
+This can be explained by the fact that each NTP sample is individually prepared and is subject to more variation in this action than compared to a machine.
+It is possible to automate the use of the NTP method, doing so may yield less variable values.
 
-## Strengths and Limitations
+Dilution and input play an important part in Ct recovery (\@ref(fig:ImportanceFactor) and \@ref(fig:ImportanceFactor2)). 
+Tools A and B are different dilution levels (1:10 and 1:1000, respectively) and the drop between Ct values is consistent with expected log loss equivalent values (\@ref(fig:P2-Tool-AB)).
+In \@ref(fig:P2-Tool-AB) we can see that as the input increases the general sample has a smaller recovery Ct value.
+Yet, with the data gathered only after the input is greater than 6800uL does the value surpass the recovery range obtained at the 1000ul input volume.
+This suggests that the NTP method is able to utilize and capture at least what is available in the sample, and perhaps more.
+
+## Limitations
+
+The use of dilution in this study is limiting as we can only estimate the genomic copies that are being picked up by the real time RT-PCR assay.
+Additional studies or re-count of the samples by use of digital droplet PCR to get an estimate count of genomic copies would yield better association and capture comparison data.
+
 RT-PCR does not distinguish between infectious and non-infectious virus. Additional work will need to be completed to better compare the recovery and estimate the true risk of SC2 in the environment.
+
 Need more replicates, but this is a time consuming processes just to get the data.
@@
 Previous studies have demonstrated capture techniques can overcome limitations of conventional studies, and if applied to environmental samples can yield better data to enlighten the field [@Gasc2018].
+
 ## Conclusions
-The use of nanoparticles does result in the recovery of Ct values in the assay.
-with the results of the analysis of data there is no significant difference between the automated extraction method and the nanoparticles.
+
+The use of NTP does result in the recovery of Ct values in the assay.
+with the results of the analysis of data there is no significant difference between the automated extraction method and the NTP
 Using some ML modeling we can see that there are some recorded variables that affect the Ct values- Dilution of the sample and Input.
 
+# Data accessibility
+
+Data and code required to complete the analysis are available from `MONICACHAN-MADA-project` Repository.
+
 # References

### 2021-12-03 reviews/peer_review_priyanka.Rmd
Peer Review P. Commit
Added rmd due to issues with pull request.

---
title: Project Review - Priyanka 
date: "`r file.mtime(knitr::current_input())`"
#bibliography: ../media/references.bib
output: 
  html_document:
    toc_depth: 3
    number_sections: true
---

# Overview

Title of project:

Name of project author(s): Monica Chan

Name of project reviewer: Priyanka Gannavarapu

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/knits into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The project's context is clearly defined, as is the rationale for investigating this subject. The introduction can have some more information added( previous research)

### Summary assessment (PICK ONE, DELETE THE OTHERS)
*some contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

 The objective is based on the data. But clear hypothesis and questions are needed to make it more clear

### Summary assessment
*question/hypotheses somewhat explained

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The data is clearly described, as is the source.

### Summary assessment
* Source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

 More explanation required for data cleaning. Detailed data wrangling would better to understand what is going on! Exploratory analysis should also be presented with more detailed explanation

### Summary assessment
* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Data analysis layout has been given well.

### Summary assessment

* Defensible but not optimal analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The flow of the results presented but needs more commentary and detailed description of what the findings are

### Summary assessment

* Results are presented ok, with room for improvement

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

Discussion section can be worked on a little. Presenting the finding and discussing the previous literature would be a solution. Checking if the results are in line or different from previous studies

### Summary assessment
* Minor parts wrong, missing or unclear

## Further comments

_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._

# Overall project content evaluation
Evaluate overall features of the project by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

The discussion portion can be worked on a little by mentioning previous research and comparing if the findings of this study are in line or is their any difference. If yes, what could the reasons be for the change in the results.

### Summary assessment
mostly clear, but some confusing parts

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

 The documentation needs more clearer explanation and should be distinct to show what information is in which folder. Needs more detailed comments in the specific files

### Summary assessment
* decently documented with some gaps

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Reproducible results

### Summary assessment
* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

Questions and hypothesis were addressed. More alternatives can be considered

### Summary assessment
* decent level of thoroughness

## Further comments

_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project.

Great project considering the situation we are in. all results reproducible and no manual intervention was needed.

### 2021-12-09 reviews/Project_Review_ESuez_to_MChan.rmd
Review merge
Moved peer-reviews into "review" folder.

--- 2021-12-03 reviews/peer_review_priyanka.Rmd
@@
----
-title: Project Review - Priyanka 
-date: "`r file.mtime(knitr::current_input())`"
-#bibliography: ../media/references.bib
-output: 
-  html_document:
-    toc_depth: 3
-    number_sections: true
----
-
 # Overview
 
-Title of project:
+Title of project:"MADA Data Analysis Project"
 
@@
 
-Name of project reviewer: Priyanka Gannavarapu
-
-# Instructions
-
-Write your comments and feedback below for each section/component of the project.
-The goal should be to help the author improve their project.
-Make comments as constructive and actionable as possible.
-You can provide both criticism and praise.
-
-For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.
-
-Make sure your final document compiles/knits into a readable, well-formatted html document.
-
-Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)
+Name of project reviewer: Ehsan Suez
 
@@
 
-The project's context is clearly defined, as is the rationale for investigating this subject. The introduction can have some more information added( previous research)
+**
+Precisely defined. I would personally want some details in introduction.**
 
 ### Summary assessment (PICK ONE, DELETE THE OTHERS)
-*some contextualization and motivation
+**some contextualization and motivation**
 
@@
 
- The objective is based on the data. But clear hypothesis and questions are needed to make it more clear
+**I think all three aims were clearly presented here**
 
 ### Summary assessment
-*question/hypotheses somewhat explained
+
+** question/hypotheses fully clear **
 
@@
 
-The data is clearly described, as is the source.
+I think the data was clearly described, as well as the source of the data.
 
 ### Summary assessment
-* Source and overall structure of data well explained
+
+** source and overall structure of data well explained **
 
@@
 
- More explanation required for data cleaning. Detailed data wrangling would better to understand what is going on! Exploratory analysis should also be presented with more detailed explanation
+I think you have done a great job with data wrangling. But I would suggest to add more comments for the readers to understand the way you were thinking while doing those particular steps.
 
 ### Summary assessment
-* some weaknesses in wrangling and exploratory component
+
+** some weaknesses in wrangling and exploratory component **
 
@@
 
-Data analysis layout has been given well.
+As per my knowledge, I think the analysis were pretty sound. And the way the analysis was presented, was great. There is always room for improvement in general, but I think you did great here.
 
 ### Summary assessment
-
-* Defensible but not optimal analysis
+* strong and reasonable analysis
 
@@
 
-The flow of the results presented but needs more commentary and detailed description of what the findings are
+The result presentation was good. Tables and figures were easy to understand.
 
@@
 
-* Results are presented ok, with room for improvement
+* results are very well presented
 
@@
 
-Discussion section can be worked on a little. Presenting the finding and discussing the previous literature would be a solution. Checking if the results are in line or different from previous studies
+I think the finding was discussed on a good level. You can add some more points there (detailing basically).
 
 ### Summary assessment
-* Minor parts wrong, missing or unclear
+
+* strong, complete and clear discussion
 
 ## Further comments
-
-_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._
 
@@
 
-The discussion portion can be worked on a little by mentioning previous research and comparing if the findings of this study are in line or is their any difference.
-If yes, what could the reasons be for the change in the results.
+We;; structured, well labeled. Some files in the analysis folder could have been named a bit better.
 
 ### Summary assessment
-mostly clear, but some confusing parts
+
+** mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders) **
 
@@
 
- The documentation needs more clearer explanation and should be distinct to show what information is in which folder. Needs more detailed comments in the specific files
+Overall well documented. But I would suggest having a clearer way of labeling files and organizing them.
 
 ### Summary assessment
+
 * decently documented with some gaps
@@
 
-Reproducible results
+I was able to rerun the whole thing.
 
@@
 
-Questions and hypothesis were addressed. More alternatives can be considered
+WRITE YOUR FEEDBACK HERE
 
 ### Summary assessment
-* decent level of thoroughness
+
+* strong level of thorougness
 
@@
 
-_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project.
-
-Great project considering the situation we are in. all results reproducible and no manual intervention was needed.
+_Add any other comments regarding the overall project here. Write anything you think can help your classmate improve their project._

### 2021-12-13 ARCHIVED/code/processing_code/TEMP/Cleaning_Reports.rmd
File Clean-up
Clean up of files

--- 2021-11-12 code/processing_code/TEMP/Cleaning_Reports.rmd
@@
 
-#create path for all csv files in directory of reports
+#create path for all csv files in directory
 csv_files<-fs::dir_ls(data_dir, # shows list of files in directory

### 2021-12-13 ARCHIVED/code/processing_code/TEMP/P2.Data.Asks.RMD
File Clean-up
Clean up of files

--- 2021-11-04 code/processing_code/P2.Data.Asks.RMD
@@
   mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
-  filter(Method=="NTP", Input=="300"|Input=="1000")%>%
+  filter(Method=="NTP", Input<=6000)%>%
   ggplot(aes(x=Dilution, y=Ct, color=Tool))+
@@
   geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
-  facet_wrap(.~Detector)+
+  facet_wrap(Input~Detector)+
   labs(title="NTP extraction comparisons with all at 300 and 1000 input levels and all tools")
@@
 NTP
-NTP extraciton comparisons wluent input levels and all tools
+NTP extraciton comparisons with eluent input levels and all tools
 ```{r}
@@
   mutate(Dilution=as.numeric(as.factor(Dilution)))%>%
-  filter(Method=="NTP", Tool!="SD", Input>=6000, Input!="CTRL")%>%
-  ggplot(aes(x=Dilution, y=Ct))+
+  filter(Method=="NTP", Input!="CTRL", Input>=6000)%>%
+  ggplot(aes(x=Dilution, y=Ct, color=Tool))+
+  geom_jitter()+
   scale_y_reverse()+
-  geom_boxplot(aes())+
+  geom_smooth(method=glm, fullrange=TRUE)+
   geom_hline(yintercept=27, linetype="longdash", color = "orange")+ #CDC confirmed threshold
-  facet_grid(Tool~Detector)+
-  labs(title="NTP extraction comparisons Eluents input levels and all tools")
+  geom_hline(yintercept=37, linetype="longdash", color = "blue")+ #General confirmed threshold
+  facet_wrap(.~Detector)+
+  labs(title="NTP extraction at Eluent input levels and all tools")
 ```

### 2021-12-13 products/manuscript/Manuscript.Rmd
File Clean-up
Clean up of files

@@
 ---
-title: "MADA Data Analysis Project"
+title: "Evaluation of "
 subtitle: ""
@@
 ---
-
 # Summary/Abstract
@@
 The objective of this study is improved outbreak response through rapid pathogen detection by evaluating a TC technology’s ability to capture SC2 from complex matrices.
+That data gathered supports that the using TC technology, nanoparticles, does extract from a sample, yet shows no difference between the currently used automated extraction method.
+Comparisons between the details of method used were also explored.
+Sample factors such as input and dilution were found to impact sample recovery and environmental data showed very light effects-- as shown by using machine learning methods.
+
+<br><br>
 
@@
 The virus is capable of human-to-human transmission via inhalation of infected aerosols.
-Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020].
+To limit the effect, patients with COVID-19 symptoms have been encouraged to test for a confirmed diagnosis one such test that has been the accepted standard is the SARS-coV-2 real-time RT-PCR reverse transcription–polymerase chain reaction (RT-PCR)[@Hanson2020].
+However, like with many tests, RT-PCR are subject to error and can produce false negative results.
+
+This Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020].
+
+CT quantification has been shown to have potential for assisting in decision-making in the management of COVID-19 patients[@Hanson2020].
+CT quantification combined with patient symptom data has shown promise when implemented with machine learning, although many studies are focused on patient information or are proof of concept [@Cai2020;@Langer2020;@Gangloff2021].
+
+The promise of these evaluations may translate well into environmental sampling and surveillance allowing for preventive action in regards to reducing risk and containing outbreak.
+<br>
 

### 2021-12-13 products/manuscript/Manuscript.Rmd
Wordings
Update descriptions in manuscript

--- 2021-11-29 products/manuscript/Manuscript.Rmd
@@
 ---
-title: "MADA Data Analysis Project"
+title: "Evaluation of sample enhancment: targeted capture with nanoparticles compared with an automated extraction method"
 subtitle: ""
@@
 
-The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants.
-Environmental reservoirs are a reported, indirect transmission route for viral pathogens [@Tanner2021;@Marcenac2021-sz].
-Although many viruses have a low infectious dose, they can be difficult to detect due to the low bio-burden in the environment.
-The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control [@Paton2021-jc].
+The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants has encouraged a continued need for better and faster testing.
+
 Culture-based methods can be labor and time intensive.
+An accepted standard for testing humans for COVID-19 is the SC2 real-time RT-PCR reverse transcription–polymerase chain reaction (RT-PCR)[@Hanson2020].
+CT quantification has been shown to have potential for assisting in decision-making in the management of COVID-19 patients[@Hanson2020].
+
+As the virus is capable of human-to-human transmission via inhalation of infected aerosols, the existence of environmental reservoirs is likely.
+Environmental reservoirs are a known indirect transmission route for viral pathogens [@Tanner2021;@Marcenac2021-sz].The ability to detect it in an environment is critical to identifying hidden reservoirs which can inform infection control [@Paton2021-jc].
+Many viruses have a low infectious dose, they can be difficult to detect due to the low bio-burden in the environment.
+
+Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020].
 The use and limitations of commercial "target capture" (TC) technologies are varied compared to molecular approaches utilizing automated extraction.
 The objective of this study is improve outbreak response through rapid pathogen detection by evaluating a TC technology's ability to capture SC2 from complex matrices.
+The TC method proposed to be tested is one based on magnetized hydro-gel particles treated with affinity dyes that can be customized to select for and bind with a multitude of analytes.
+A retail kit of these particles, NanoTrap Particles, are commercially avalible and advertised as being specially capable of "capturing" the SC2 virus from various sample types such as clinical isolates to waste water better than processing a sample without the use of the particle.
 
-The virus is capable of human-to-human transmission via inhalation of infected aerosols.
-Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020].
+Using RT-PCR we will establish that the TC method, NanoTrap Particles, perform.
+Then the results will be compared with a standardized automated preparation method.
+By measuring the Ct values of the  resulting methods we can determine the sensitivity of these extraction methods to capture addition or retain the genomic copies from a sample.
+The data set will then be broken down and analyzed specifically with surface sampling capture to see if it is possible to recover from the environment with the use of a specialized sampling tool for molecular work.
+
+CT quantification combined with patient symptom data has shown promise when implemented with machine learning, although many studies are focused on patient information or are proof of concept [@Cai2020;@Langer2020;@Gangloff2021].
+Additional exploration of the data with some machine learning techniques for any significant importance to Ct value recovery will be performed.
+<br>
 
@@
 
-Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR.
+Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; NanoTrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by RT-PCR.
 [@Wittwer2012-vt;@Shah2021-mb;@Barclay2020].
@@
 
-Data was cleaned and reformatted to relevant factors and numeric data types.
+Data was checked for missing data points (aside from Ct Values) and reformatted factors were mutated to the correct factor or numeric data types.
 During cleaning the data was split and saved in three forms- All (compilation of all recovery studies- P1 and P2);P1 (only data from spiked studies); and P2 (only data from surface sampling studies).
+
 P1 and P2 were analyzed together due to constant serial dilution controls.
 P2 data were expanded on due to the interests of the study focusing on surface recovery.
+
+Cleaning scripts are found in the `ProcessingScript.Rmd` file in the code folder.
 
@@
 
-For Aim 2, data was limited to P2 data from this point onwards to focus on surface recovery of genomic material and limit variability from P1 spiked studies.
+For Aim 2, data was limited to P2 data from this point onward to focus on surface recovery of genomic material and limit variability from P1 spiked studies.
 An exploration in how Tool and Methods differed.
-Additional exploration occured in looking specifically at the NTP Method and its results as there were additional variables in input.
+Additional exploration occurred in looking specifically at the NTP Method and its results as there were additional variables in input.
 
@@
 Overall the Methods appear to not be too different from one another at higher dilution levels, but we also see that as the Dilution level decreases the the NTP method registers a lower Ct value than the CDC method and begins to differentiate.
+_Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_.
 
@@
 The divergence of CDC and NTP methods is still seen slightly with CDC recovering less at higher dilution levels and switching places with NTP as the dilution increases.
+_Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_
 
@@
 A log drop is ~3Ct values we can see from the graphic that it trends as predicted.
-Between methods at an input level of 300ul, there's a lot more variability in the NTP method than the CDC method,.
+Between methods at an input level of 300µl, there's a lot more variability in the NTP method than the CDC method,.
 Probes see little to no difference in Ct Values recovered.
+_Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_
 
-### Phase 2 Comparing Methods with standard input of 300uL
+### Phase 2 Comparing Methods with standard input of 300µL
 
@@
 Might be significant difference at the -4 dilution.
+_Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_
 
-### Phase 2 Nanoparticles only - All tools at 300 and 1000ul input levels
+### Phase 2 Nanoparticles only - All tools at 300 and 1000µl input levels
 
@@
 Still no difference between Probe types.
+_Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_
 
@@
 In \@ref(fig:P2-Tool-AB) we can see that as the input increases the general sample has a smaller recovery Ct value.
-Yet, with the data gathered only after the input is greater than 6800uL does the value surpass the recovery range obtained at the 1000ul input volume.
+Yet, with the data gathered only after the input is greater than 6800µL does the value surpass the recovery range obtained at the 1000µl input volume.
 This suggests that the NTP method is able to utilize and capture at least what is available in the sample, and perhaps more.
@@
 
-Data and code required to complete the analysis are available from `MONICACHAN-MADA-project` Repository.
+Data and code required to complete the analysis are available from `MONICACHAN-MADA-project` [Repository](https://github.com/MoChan2021/MONICACHAN-MADA-project) .
 

### 2021-12-13 products/manuscript/Manuscript.Rmd
Final Commit

@@
 Sample factors such as input and dilution were found to impact sample recovery and environmental data showed very light effects-- as shown by using machine learning methods.
-
-<br><br>
+<br>
 
@@
 The data set will then be broken down and analyzed specifically with surface sampling capture to see if it is possible to recover from the environment with the use of a specialized sampling tool for molecular work.
-
-CT quantification combined with patient symptom data has shown promise when implemented with machine learning, although many studies are focused on patient information or are proof of concept [@Cai2020;@Langer2020;@Gangloff2021].
-Additional exploration of the data with some machine learning techniques for any significant importance to Ct value recovery will be performed.
 <br>
@@
 If dilutions are above -1.5, the input of sample and method may influence the recovered Ct values.
+<br>
 
@@
 
-RT-PCR does not distinguish between infectious and non-infectious virus. Additional work will need to be completed to better compare the recovery and estimate the true risk of SC2 in the environment.
+RT-PCR does not distinguish between infectious and non-infectious virus.
+Additional work will need to be completed to better compare the recovery and estimate the true risk of SC2 in the environment.
+Ct quantification combined with patient symptom data has shown promise when implemented with machine learning [@Cai2020;@Langer2020;@Gangloff2021].
+A number of studies are focused on patient information or are proof of concept [@Cai2020;@Langer2020;@Gangloff2021].
+Additional exploration of environmental recovery data with similar machine learning techniques could enhance the ability to predict or back calculate true source values in the environment.
 
-Need more replicates, but this is a time consuming processes just to get the data.
+The data in general need more replicates, but this is a time consuming processes just to get the data.
+This could be remedied by additional dummy variables, but additional replicates would also help with the dummy generation.
 
-Additional materials need to be considered [@Jansson2020-cg].
+Additional factors such as materials and sample area need to be considered [@Jansson2020-cg].
+Previous work with bacteria have shown that there are affects on surface area and surface tool on total recovery [@Tanner2021;@Marcenac2021-sz].
+
 Previous studies have demonstrated capture techniques can overcome limitations of conventional studies, and if applied to environmental samples can yield better data to enlighten the field [@Gasc2018].
@@
 Using some ML modeling we can see that there are some recorded variables that affect the Ct values- Dilution of the sample and Input.
+<br>
 
